---
title: "PUBPOL 5390 Homework 2"
author: Thelonious Goerz
date: "Draft date: `r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document:
    number_sections: yes
linestretch: 1.5
fig_width: 5
fig_height: 2.5
csl: "/Users/theloniousgoerz/Academic/Professional/Resources/Citations/styles-master/american-sociological-association.csl"
editor_options:
  chunk_output_type: inline
geometry: "left=1in,right=1in,top=1in,bottom=1in"
number-sections: true
keep_tex: true
link-citations: yes
linkcolor: blue
font_size: 12pt
header-includes: 
  - \usepackage{framed} 
  - \usepackage{amsmath}
  - \usepackage{tcolorbox} 
  - \usepackage{pdflscape}
  - \newcommand{\blandscape}{\begin{landscape}}
  - \newcommand{\elandscape}{\end{landscape}}
  - \usepackage{multirow} 
  - \usepackage{xcolor} 
  - \usepackage{tabularray}
  - \def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}} 
  - \usepackage{setspace} 
  - \usepackage{threeparttable}
  - \usepackage{booktabs}
  - \usepackage{siunitx}
  - \usepackage{tikz}
  - \usepackage{tabularray}
  - \usepackage{float}
  - \usepackage{graphicx}
  - \usepackage{rotating}
  - \usepackage[normalem]{ulem}
  - \UseTblrLibrary{booktabs}
  - \UseTblrLibrary{siunitx}
---

```{r setup, include=TRUE, results="hide", echo = FALSE, message = F}

 rm(list = objects())
 # THIS IS THE WIDTH OF THE CODE                                                ###
 knitr::opts_chunk$set(echo = F,
                       resuls = T, message = F, warning = F)
 
 # knitr::opts_chunk$set(fig.width = unit(16,"cm"), fig.height = unit(9,"cm"))
 
```

\doublespacing
\fontsize{12}{22}
\setstretch{2}
\baselineskip=29pt

```{r Load Packages}
rm(list = ls())
### Packages ###
library(readr)
library(tidyverse)
library(fixest)
library(modelsummary)
library(haven)
library(magrittr)
library(splines)
library(glmnet)
library(KernSmooth)
library(boot)
library(tidymodels)
library(here)
library(KernSmooth)

```

```{r Data}
# Problem 1 
Gsmall = read_csv(here("Data","births_for_class_Gsmall.csv"))
H = read_csv(here("Data","births_for_class_H.csv"))
G = read_csv(here("Data","births_for_class_G.csv"))
ahs =   read.csv(here("Data","ahs2011n.csv"),quote="\"'")
```

# Question 1

```{r Feature Set}
set1 = c(
    "gestation_weeks",
    "mom_age",
    "mom_race_hispanic_status",
    "mom_educ")

set2 = c(    
    "mom_age",
    "mom_race_hispanic_status",
    "mom_educ",
    "gestation_weeks",
    "time",
    "in_hospital",
    "gestation_weeks"
    )

set3 = c(    
    "mom_age",
    "mom_race_hispanic_status",
    "mom_educ",
    "gestation_weeks",
    "time",
    "in_hospital",
    "gestation_weeks",
    "prior_births_living",
    "plurality",
    "residence_status",
    "mom_bmi",
    "dad_age"
    )

FS_1 = paste(
set1,
collapse = "+"
)

FS_2 = paste(
set2,
collapse = "+"
)

FS_3 = paste(
set3,
collapse = "+"
)
```

```{r}
# Label Train and Test Set 
training = Gsmall
testing = H
```

```{r OLS}
ols_function = function(outcome,
                        training,
                        testing,
                        FS_recipe
                        ){
## SD Y ------------------------------
is_sd_y = sd(training[[outcome]])
os_sd_y = sd(testing[[outcome]])
## Define Model ------------------------------
ols = linear_reg(
  mode = "regression",
  engine = "lm"
) 
## Define Workflow ------------------------------
ols_workflow = 
 workflow() %>% 
  add_model(ols) %>% 
  add_recipe(FS_recipe)

## Fit Model -------------------------------------
fit_ols =
  ols_workflow %>% 
  fit(data = training)

## Obtain Predictions -----------------------------
ols_oos = predict(fit_ols,testing)
  
# Calculate RMSE ---------------------------------

## Combine Train-Test predictions and Outcomes into a DF 
ols_gof_train_df = data.frame(
  estimate = as.numeric(predict(fit_ols,training)$.pred),
  truth = unlist(fit_ols$pre$mold$outcomes[[outcome]]))

ols_gof_test_df = data.frame(
  estimate = as.numeric(ols_oos$.pred),
  truth = as.numeric(testing[[outcome]]))

## Calculate RMSE ------------------------------
ols_train_rmse = rmse(
     data = ols_gof_train_df,
    truth =  truth,
    estimate =  estimate)

ols_test_rmse = rmse(
     data = ols_gof_test_df,
     truth = truth,
     estimate = estimate)

oos_r2 = rsq(data = ols_gof_test_df,
     truth = truth,
     estimate = estimate)$.estimate %>% as.numeric()
is_r2 = rsq(data = ols_gof_train_df,
     truth = truth,
     estimate = estimate)$.estimate %>% as.numeric()
# Quitiles of price  ------------------------------
quintiles = ols_gof_test_df %>%
 mutate(quintile = cut(
      truth,
      breaks = quantile(truth, probs = seq(0, 1, 0.20), na.rm = TRUE),
      include.lowest = TRUE,
      labels = paste0("P", 1:5)
    )) %>% 
  # summarise by metric 
  group_by(quintile) %>% 
  summarise(mu_lv_ols = mean(truth),
            mu_p_ols = mean(estimate),
            ols_diff = (mu_lv_ols - mu_p_ols)
            ) %>%
  select(quintile,ols_diff) %>% 
  pivot_wider(names_from = "quintile", values_from = "ols_diff")

# Return Items ------------------------------
return(
  cbind(data.frame(
    Model = "OLS",
    in_sample_sd_y = is_sd_y,
    oo_sample_sd_y = os_sd_y,
    in_sample_rmse = ols_train_rmse$.estimate,
    oo_sample_rmse = ols_test_rmse$.estimate,
    n_predictors = ncol(fit_ols$pre$mold$predictors),
    is_r2 = is_r2,
    oos_r2 = oos_r2
  ),
  # quintiles
  quintiles
  )
)
}
```

```{r LASSO}
regularize_function = function(outcome,
                        training,
                        testing,
                        FS_recipe,
                        type,
                        n_folds,
                        grid_max
                        ){
## SD Y ------------------------------
is_sd_y = sd(training[[outcome]])
os_sd_y = sd(testing[[outcome]])
## Tune Parameters ------------------------------
## Pick Model

## Define folds 
folds = vfold_cv(training,v = n_folds) 

if (type == "LASSO"){
model_tune = linear_reg(
  mode = "regression",
  engine = "glmnet",
  mixture = 1, 
  penalty = tune()
) 
# Cross-Validate
cv_lambda = tune_grid(model_tune, 
                      FS_recipe,
                      resamples = folds, 
                      tune = penalty,
                      grid = expand.grid(penalty = 0:grid_max))

# Pick lambda # 
best_penalty = as.numeric(show_best(cv_lambda)[1,"penalty"])
best_mixture = 1

} else if (type == "Ridge"){
model_tune = linear_reg(
  mode = "regression",
  engine = "glmnet",
  mixture = 0, 
  penalty = tune()
) 
# Cross-Validate
cv_lambda = tune_grid(model_tune, 
                      FS_recipe,
                      resamples = folds, 
                      tune = penalty,
                      grid = expand.grid(penalty = 0:grid_max))

# Pick lambda # 
best_penalty = as.numeric(show_best(cv_lambda)[1,"penalty"])
best_mixture = 0

  
} else if (type == "Elastic"){
model_tune = linear_reg(
  mode = "regression",
  engine = "glmnet",
  mixture = tune(), 
  penalty = tune()
) 
# Cross-Validate 

cv_pick = tune_grid(model_tune, 
                      FS_recipe,
                      resamples = folds, 
                      tune = c(penalty,mixture),
                      grid = expand.grid(penalty = 0:grid_max,
                                         mixture = c(.25,.5,.75)))

# Pick lambda # 
best_penalty = as.numeric(show_best(cv_pick)[1,"penalty"])
best_mixture = as.numeric(show_best(cv_pick)[1,"mixture"])
}

# ------------------------------------------------------------
# Declare Model # 
regularized_model = linear_reg(
  mode = "regression",
  engine = "glmnet",
  mixture = best_mixture, 
  penalty = best_penalty
) 
## Define Workflow ------------------------------
rglr_workflow = 
 workflow() %>% 
  add_model(regularized_model) %>% 
  add_recipe(FS_recipe)

## Fit Model -------------------------------------
fit_rglr =
  rglr_workflow %>% 
  fit(data = training)

## Obtain Predictions -----------------------------
rglr_oos = predict(fit_rglr,testing)

tidy_coefficients <- fit_rglr %>%
      extract_fit_parsnip() %>%
      tidy()
    # Filter for non-zero coefficients (selected predictors)
    n_terms <- tidy_coefficients %>%
      filter(estimate != 0) %>%
      pull(term) %>% 
      length() %>% 
      as.numeric()
  
# Calculate GOF ---------------------------------
## Combine Train-Test predictions and Outcomes into a DF 
rglr_gof_train_df = data.frame(
  estimate = as.numeric(predict(fit_rglr,training)$.pred),
  truth = unlist(fit_rglr$pre$mold$outcomes[[outcome]]))

rglr_gof_test_df = data.frame(
  estimate = as.numeric(rglr_oos$.pred),
  truth = as.numeric(testing[[outcome]]))

## Calculate GOF ------------------------------
rglr_train_rmse = rmse(
     data = rglr_gof_train_df,
     truth,
     estimate)

rglr_test_rmse = rmse(
     data = rglr_gof_test_df,
     truth,
     estimate)

oos_r2 = rsq(data = rglr_gof_test_df,
     truth = truth,
     estimate = estimate)$.estimate %>% as.numeric()
is_r2 = rsq(data = rglr_gof_train_df,
     truth = truth,
     estimate = estimate)$.estimate %>% as.numeric()

# Quintiles ------------------------------
quintiles = rglr_gof_test_df %>%
 mutate(quintile = cut(
      truth,
      breaks = quantile(truth, probs = seq(0, 1, 0.20), na.rm = TRUE),
      include.lowest = TRUE,
      labels = paste0("P", 1:5)
    )) %>% 
  # summarise by metric 
  group_by(quintile) %>% 
  summarise(mu_lv = mean(truth),
            mu_p = mean(estimate),
            diff = (mu_lv - mu_p)
            ) %>%
  select(quintile,diff) %>% 
  pivot_wider(names_from = "quintile", values_from = "diff")
# Return Items ------------------------------
 return(
   cbind(data.frame(
     Model = {{type}},
     in_sample_sd_y = is_sd_y,
     oo_sample_sd_y = os_sd_y,
     in_sample_rmse = rglr_train_rmse$.estimate,
     oo_sample_rmse = rglr_test_rmse$.estimate,
     n_predictors = n_terms,
     mixture = best_mixture,
     penalty = best_penalty,
     is_r2 = is_r2,
     oos_r2 = oos_r2
   ),
   quintiles
   )
 ) 
}

```


```{r Run OLS For Each Feature Set}
## Create Recipes 
FS_recipe_1 = recipe(as.formula(paste0("birth_weight","~",FS_1)), 
         data = training) %>%
         step_zv()

FS_recipe_2 = recipe(as.formula(paste0("birth_weight","~",FS_2)), 
         data = training) %>%
  step_mutate(
    across(c(
    mom_race_hispanic_status,
    mom_educ,
    gestation_weeks
    ), 
    as.factor)
  ) %>% 
    step_dummy(all_nominal_predictors(),
             one_hot = T) %>% 
  step_zv(all_predictors())

FS_recipe_3 = recipe(as.formula(paste0("birth_weight","~",FS_3)), 
         data = training) %>%
  step_mutate(
    across(c(
    mom_race_hispanic_status,
    mom_educ,
    gestation_weeks,
    time,
    ), 
    as.factor)
  ) %>% 
    step_dummy(all_nominal_predictors(),
                one_hot = T) %>% 
  step_spline_b(mom_age,
                deg_free = 6,
                degree = 3,
                keep_original_cols = T) %>%
  step_zv(all_predictors())

# Run OLS Function on feature sets 
ols_fs_1 = ols_function("birth_weight",
    FS_recipe = FS_recipe_1,
    training = training,
    testing = testing)

ols_fs_2 = ols_function("birth_weight",
    FS_recipe = FS_recipe_2,
    training = training,
    testing = testing) 

ols_fs_3 = ols_function("birth_weight",
    FS_recipe = FS_recipe_3,
    training = training,
    testing = testing) 
```


```{r Run Regularized Regressions, cache = T}
set.seed(9999)
# Run LASSO Models ------------------------------------- # 
# 
# -------------------------------------------------------#

lasso_FS_1 = regularize_function(outcome = "birth_weight",
                        training = training,
                        testing =  testing,
                        FS_recipe = FS_recipe_1,
                        type = "LASSO",
                        n_folds = 10,
                        grid_max = 30
                        )

lasso_FS_2 = regularize_function(outcome = "birth_weight",
                        training = training,
                        testing =  testing,
                        FS_recipe = FS_recipe_2,
                        type = "LASSO",
                        n_folds = 10,
                        grid_max = 30
                        )

lasso_FS_3 = regularize_function(outcome = "birth_weight",
                        training = training,
                        testing =  testing,
                        FS_recipe = FS_recipe_3,
                        type = "LASSO",
                        n_folds = 10,
                        grid_max = 30
                        )

set.seed(9999)
# Run RIDGE Models ------------------------------------- # 
# 
# -------------------------------------------------------#

ridge_FS_1 = regularize_function(outcome = "birth_weight",
                        training = training,
                        testing =  testing,
                        FS_recipe = FS_recipe_1,
                        type = "Ridge",
                        n_folds = 10,
                        grid_max = 50
                        )

ridge_FS_2 = regularize_function(outcome = "birth_weight",
                        training = training,
                        testing =  testing,
                        FS_recipe = FS_recipe_2,
                        type = "Ridge",
                        n_folds = 10,
                        grid_max = 50
                        )

ridge_FS_3 = regularize_function(outcome = "birth_weight",
                        training = training,
                        testing =  testing,
                        FS_recipe = FS_recipe_3,
                        type = "Ridge",
                        n_folds = 10,
                        grid_max = 50
                        )

# Run Elastic Net Models ------------------------------------- # 
# 
# -------------------------------------------------------------#

elastic_FS_1 = regularize_function(outcome = "birth_weight",
                        training = training,
                        testing =  testing,
                        FS_recipe = FS_recipe_1,
                        type = "Elastic",
                        n_folds = 10,
                        grid_max = 50
                        )

elastic_FS_2 = regularize_function(outcome = "birth_weight",
                        training = training,
                        testing =  testing,
                        FS_recipe = FS_recipe_2,
                        type = "Elastic",
                        n_folds = 10,
                        grid_max = 50
                        )

elastic_FS_3 = regularize_function(outcome = "birth_weight",
                        training = training,
                        testing =  testing,
                        FS_recipe = FS_recipe_3,
                        type = "Elastic",
                        n_folds = 10,
                        grid_max = 50
                        )


ols_fs_1 %<>% mutate(mixture = NA,penalty = NA)
ols_fs_2 %<>% mutate(mixture = NA,penalty = NA)
ols_fs_3 %<>% mutate(mixture = NA,penalty = NA)

# Look at Models

rbind(ols_fs_1 %>% select(Model, in_sample_sd_y, oo_sample_sd_y, is_rmse = in_sample_rmse,     oos_rmse = oo_sample_rmse, n_predictors,alpha= mixture,lambda =penalty) ,
      ols_fs_2 %>% select(Model, in_sample_sd_y, oo_sample_sd_y, is_rmse = in_sample_rmse,     oos_rmse = oo_sample_rmse, n_predictors,alpha= mixture,lambda =penalty) ,
      ols_fs_3 %>% select(Model, in_sample_sd_y, oo_sample_sd_y, is_rmse = in_sample_rmse,     oos_rmse = oo_sample_rmse, n_predictors,alpha= mixture,lambda =penalty) ,
      lasso_FS_1 %>% select(Model, in_sample_sd_y, oo_sample_sd_y, is_rmse = in_sample_rmse,   oos_rmse = oo_sample_rmse, n_predictors,alpha= mixture,lambda =penalty) ,
      lasso_FS_2 %>% select(Model, in_sample_sd_y, oo_sample_sd_y, is_rmse = in_sample_rmse,   oos_rmse = oo_sample_rmse, n_predictors,alpha= mixture,lambda =penalty) ,
      lasso_FS_3 %>% select(Model, in_sample_sd_y, oo_sample_sd_y, is_rmse = in_sample_rmse,   oos_rmse = oo_sample_rmse, n_predictors,alpha= mixture,lambda =penalty) ,
      ridge_FS_1 %>% select(Model, in_sample_sd_y, oo_sample_sd_y, is_rmse = in_sample_rmse,   oos_rmse = oo_sample_rmse, n_predictors,alpha= mixture,lambda =penalty) ,
      ridge_FS_2 %>% select(Model, in_sample_sd_y, oo_sample_sd_y, is_rmse = in_sample_rmse,   oos_rmse = oo_sample_rmse, n_predictors,alpha= mixture,lambda =penalty) ,
      ridge_FS_3 %>% select(Model, in_sample_sd_y, oo_sample_sd_y, is_rmse = in_sample_rmse,   oos_rmse = oo_sample_rmse, n_predictors,alpha= mixture,lambda =penalty) ,
      elastic_FS_1 %>% select(Model, in_sample_sd_y, oo_sample_sd_y, is_rmse = in_sample_rmse, oos_rmse = oo_sample_rmse, n_predictors,alpha= mixture,lambda =penalty) ,
      elastic_FS_2 %>% select(Model, in_sample_sd_y, oo_sample_sd_y, is_rmse = in_sample_rmse, oos_rmse = oo_sample_rmse, n_predictors,alpha= mixture,lambda =penalty) ,
      elastic_FS_3 %>% select(Model, in_sample_sd_y, oo_sample_sd_y, is_rmse = in_sample_rmse, oos_rmse = oo_sample_rmse, n_predictors,alpha= mixture,lambda =penalty)
      ) %>% 
  datasummary_df(
    title = "Comparison of Predictive Performance Across Models and Feature Sets"
  )
```


### Cross-Validation 

Within the tidymodels framework (the R analogue to scikit-learn), there are no defaults for cross-validation. Instead, I wrote my own function a la Mullinathan and Spiess, to run the models in an automated way. These functions allow the grid size for parameters to be searched over and the number of folds for cross-validation. For all models, I set the number of folds to 10, the grid size to 30 or 50 depending on whether the penalty is L1 or L2. The goodness of fit measure that I use is RMSE. 

### Model Performance 

I use three different covariate sets that layer controls in each model and change the functional form. The firs model includes linear forms of relevant predictors (e.g., mother's age and gestation). The second and third incorporate non-linear terms as well as expanded covariates. The majority of the improvement in predictive performance in richer models is obtained from gestation weeks. 

Overall, the results, in Table 1, indicate that the best model in terms of predictive performance is a LASSO regression that selects 30 predictors out of a candidate set of 39 and achieves an out-of-sample RMSE of $437$. Across ridge and elastic net, the RMSE is slightly worse but not by that much. In general, each of the second feature set models achieves similar results. The third feature set, which is the richest model, performs the worst across models. Though, the LASSO model shrinks the feature set from 542 to 94, which reduces the RMSE. Even with shrinkage, it still performs poorly. 

# Question 2

The second table describes the results of Replication and Kitchen Sink Regressions for OLS, LASSO, and Elastic Net. Across models, the replication and kitchen sink regressions do about the same. The best performing model is the Elastic net, but it is only by a small margin (e.g. 44 v 43 R2). By House Price Quintile (P1-P5) each of the kitchen sink models achieve large performance improvements in the 3rd quintile for all models. 

```{r House Price Data Cleaning, echo = F, results='hide'}

# Data Cleaning # 
# Adapted from Mullinathan and Spiess #

ahs[, sapply(ahs, is.factor)] <- lapply(ahs[, sapply(ahs, is.factor)], as.factor)

# Make some factors factors, numeric numeric
ahs$CELLAR <-  as.factor(ahs$CELLAR)
ahs$FPLWK <-   as.factor(ahs$FPLWK)
ahs$TYPE <-    as.factor(ahs$TYPE)
ahs$NUMTLT <-  as.numeric(as.character(ahs$NUMTLT))
ahs$NUMCOLD <- as.numeric(as.character(ahs$NUMCOLD))
ahs$BUSPER <-  as.numeric(as.character(ahs$BUSPER))
ahs$NUMDRY <-  as.numeric(as.character(ahs$NUMDRY))
ahs$NUMSEW <-  as.numeric(as.character(ahs$NUMSEW))
ahs$VALUE <-   as.numeric(ahs$VALUE)

# Delete rows with zero/missing value, and get the log value
ahs <- ahs[ahs$VALUE > 0,]

ahs$LOGVALUE <- log(ahs$VALUE)

# Choose variables of interest
allvars <- names(ahs)
charvar <- intersect(allvars,c("ACCESS", "ACCESSB", "AGERES", "AIR", "AIRSYS", "ANYCAR", "ANYPNT", "ANYRUG", "APTFL", "ASSTSERV", "ATBSUN", "BATHS", "BDCARP", "BEDRMS", "BEDX", "BUILT", "BURNER", "BUSIN", "BUSPER", "CELLAR", "CLIMB", "CONDO", "COOK", "DENS", "DINING", "DIRAC", "DISH", "DISPL", "DRSHOP", "DRY", "DSTEAM", "DSTOVE", "EAIRC", "EBAR", "ECNTAIR", "EDISHWR", "EDRYER", "EFRIDGE", "EHEATUT", "ELEV", "ENOEAPP", "EOTEAPP", "ETRSHCP", "EWASHR", "EXCLUS", "FAMRM", "FLOODPLN", "FLOORS", "FLOT", "FPINS", "FPLWK", "FRPL", "FRPLI", "FRSTOC", "GARAGE", "HALB", "HALFB", "HOTPIP", "INCP", "KEXCLU", "KITCH", "KITCHEN", "LAT70S", "LAUNDY", "LIVING", "LOT", "LVCARP", "MOBILTYP", "MOPERM", "MOSALL", "MOSAPR", "MOSAUG", "MOSDEC", "MOSFEB", "MOSJAN", "MOSJUL", "MOSJUN", "MOSMAR", "MOSMAY", "MOSNOV", "MOSOCT", "MOSSEP", "NEWC", "NOOTHRM", "NUMAIR", "NUNIT2", "NUNITS", "OARSYS", "OBEDRM", "ODIN", "ODIRAC", "OKITCH", "OLIVIN", "OOTHRM", "OTBUP", "OTHFN", "OTHLQ", "OTHLQ1", "OTHLQ2", "OTHRUN", "OVEN", "OWNHERE", "OWNLOT", "PERSERV", "PHONE", "PLUGS", "PORCH", "PREOCC", "PUBSEW", "RECRM", "REFR", "ROOMS", "ROOMSA", "ROOMSB", "SEWDIS", "SEWDISTP", "SEWDUS", "SHARAT", "SHARFR", "SINHV", "SINK", "SINVV", "SOTHV", "SRECV", "SRENTV", "SSELV", "STEAM", "STORG", "TELAV", "TELHH", "TENURE", "TIMSHR", "TRASH", "TUB", "TYPE", "UNITSF", "VACANC2", "VACANC2B", "VACANCY", "WASH", "WATER", "WATERD", "WELLDIS", "WELLDIS2", "WELDUS", "WELL", "WFPROP", "WHNGET", "WINTERELSP", "WINTERKESP", "WINTERNONE", "WINTEROVEN", "WINTERWOOD", "YRRND"))
qualvar <- intersect(allvars,c("BADSTEP", "BDSTPQ", "BLDMNT", "BLEAK", "BSINK", "CRACKS", "DFIRE", "DISAS", "DORREP", "EBOARD", "EBROKE", "ECRUMB", "EGOOD", "EHOLER", "ELEVWK", "EMISSR", "EMISSW", "EROACH", "ESAGR", "ESLOPW", "EVROD", "EXPOSE", "FREEZE", "GRDMNT", "HOLES", "HOWH", "IFBLOW", "IFCOLD", "IFDRY", "IFSEW", "IFTLT", "ILEAK", "IMAINT", "LEAK", "LTSOK", "LTSOK1", "LTSOK2", "M12ROACH", "M12ROD", "M3ROD", "MAJR1", "MAJR2", "MAJR3", "MICE", "MINR1", "MINR2", "MINR3", "NLEAK1", "NLEAK2", "NOTSUR", "NOWIRE", "NUMBLOW", "NUMCOLD", "NUMDRY", "NUMSEW", "NUMTLT", "OMAINT", "OTHCLD", "OTLEAK", "PILEAK", "PLEAK", "PLUMB", "RAILOK", "RAILOK1", "RAILOK2", "RATFREQ", "RATS", "RLEAK", "ROACHFRQ", "TALWIR", "TASB", "TOILET", "TRADON", "TREP", "TWATER", "WATERS", "WHYCD1", "WHYCD2", "WHYCD3", "WHYCD4", "WHYCD5", "WLEAK", "WTRHRL", "ZADEQ"))
geovar <- c("REGION"
            #"METRO","METRO3","NATLFLAG"
            )
keepvars <- unique(c("LOGVALUE",geovar,charvar,qualvar))


# Limit to owner-occupied units with non-missing value and square-footage from metropolitan sample
ahs = 
  ahs[#ahs$TENURE == 1,
          ahs$VACANCY == -6,
           # & ahs$NATLFLAG == 2, remove bc not using metro sample
                  keepvars]


# Collapse very rare factor levels by hand so that don't have to delete full variable
ahs$SEWDIS[as.numeric(as.character(ahs$SEWDIS)) > 0] <- 1
ahs$WATER[as.numeric(as.character(ahs$WATER)) == 6] <- 7
ahs$ELEV[as.numeric(as.character(ahs$ELEV)) == 3] <- -6
ahs$ELEVWK[as.numeric(as.character(ahs$ELEVWK)) == 2] <- -6

# Take samples
set.seed(20170313)

ahs$holdout <- as.logical(1:nrow(ahs) %in% sample(nrow(ahs), nrow(ahs) - 10000))

# Create folds for lasso barcode plot
ahs$lassofolds <- as.factor(ceiling(10 * sample(nrow(ahs)) / nrow(ahs)))

## fix ## 
# Introduce missingness dummies
# Fix NA values 

replace_na_numeric <- function(df, na_value = -1) {
  df %>%
    mutate(across(where(is.numeric), ~ replace_na(., na_value)))
}

ahs = replace_na_numeric(ahs,na_value = -1)

#ahs$holdout = as.logica(ahs$holdout)

withmiss <- c()
for (contvar in names(ahs)[(!sapply(ahs, is.factor)) & sapply(ahs, is.numeric)]) {
  if (sum(ahs[[contvar]] < 0, na.rm = TRUE) > 0) {
    
    # make MISS indicator
    ahs[[paste0(contvar, "MISS")]] <- factor(pmin(ahs[[contvar]], 0))
    
    # replace negatives with mean of positive non-holdout values
     ahs[[contvar]][ahs[[contvar]] < 0] <- mean(ahs[[contvar]][!ahs$holdout & ahs[[contvar]] > 0],
      na.rm = TRUE)
    withmiss <- c(withmiss, contvar)
  }
}

# Collapse missingness that does not have enough variation
for(varname in names(ahs)[sapply(ahs, is.factor)]) {
  thiscol <- as.numeric(as.character(ahs[,varname]))
  if(min(table(thiscol[thiscol < 0]),na.rm =T) < 30) {
    ahs[,varname] <- as.numeric(as.character(ahs[,varname]))
    ahs[thiscol < 0,varname] <- -1
    ahs[,varname] <- as.factor(ahs[,varname])
  }
}

# Delete variables that are constant on full training set
for(varname in setdiff(names(ahs), "holdout")) {
  if(length(unique(ahs[!ahs$holdout, varname])) == 1) {
    ahs <- ahs[, !(names(ahs) %in% paste0(varname, c("", "IMPUTED")))]
  #  print(paste0("Deleted ",varname))
  }}

# Delete variables that have new levels on holdout
for(varname in names(ahs)[sapply(ahs, is.factor)]) {
  if((! varname %in% c("holdout","lassofolds")) & length(setdiff(unique(ahs[ahs$holdout, varname]),
                                                                 unique(ahs[!ahs$holdout, varname]))) > 0) {
    print(paste0(varname," misses ",setdiff(unique(ahs[ahs$holdout, varname]),
                                            unique(ahs[!ahs$holdout, varname]))))
    ahs <- ahs[, !(names(ahs) == varname)]
  }
}

# Flatten factors
ahs <- droplevels(ahs)

downtoearth <- function(RHS,survivors) {
  keepthem <- intersect(c(RHS,paste0(RHS,"MISS")),survivors)
  return(keepthem)
}
vars <- downtoearth(c(charvar,qualvar,geovar),names(ahs))

getformula <- function(varlist,LHS="LOGVALUE") {
  return(paste(LHS, paste(varlist,collapse=" + "), sep = " ~ "))
}

# Fix character Factor 
miss_factor_recode = function(df) {
  factor_cols <- names(df)[sapply(df, is.factor)]
  
  for (col in factor_cols) {
    # Convert factor to character first to preserve unusual levels
    levels_vec <- as.character(df[[col]])
    
    # Replace NA with a string so dummy creation works
    levels_vec[is.na(levels_vec)] <- "NA"
    
    # Create two dummy variables
    df[[paste0(col, "_dum1")]] <- as.integer(levels_vec %in% unique(levels_vec)[1])
    df[[paste0(col, "_dum2")]] <- as.integer(levels_vec %in% unique(levels_vec)[2])
    
    # Optionally remove original factor
    # df[[col]] <- NULL
  }
  
  return(df)
}

ahs = miss_factor_recode(ahs) %>% select(!is.factor, -EXCLUS)

ahs %<>% na.omit %>% droplevels()

```

```{r}
# House Price Data Processing
ahs_training = ahs[ahs$holdout == FALSE,]
ahs_testing =  ahs[ahs$holdout == TRUE,]
# OLS -------------------- -------------------------------------
## Replication Covariates -------------------------------------
# Create feature set

ahs_recipe_1 = recipe(LOGVALUE~.,
                data = ahs_training) %>%
  step_mutate(across(c(
    where(is.character)
    ), 
    as.factor)) %>%
  step_dummy(all_nominal_predictors(),one_hot = T) %>% 
         step_zv(all_predictors())

# Run OLS 
ahs_ols_fs_1 = ols_function("LOGVALUE",
    FS_recipe = ahs_recipe_1,
    training = ahs_training,
    testing =  ahs_testing)
## Updated Covariates -------------------------------------
ahs_recipe_2 = recipe(LOGVALUE~.,
                data = ahs_training) %>%
  step_discretize(c(BUILT,WHNGET,ROOMS,LOT), num_breaks = 10) %>%
  step_mutate(across(c(
    where(is.character),
    REGION,
    BEDRMS,
    FLOORS,
    DRSHOP,
    GARAGE,
    AIR
    ), 
    as.factor)) %>%
  step_dummy(all_nominal_predictors(),one_hot = T) %>% 
         step_zv(all_predictors())

ahs_ols_fs_2 = ols_function("LOGVALUE",
    FS_recipe = ahs_recipe_2,
    training = ahs_training,
    testing =  ahs_testing)
# LASSO -------------------- -------------------------------------
set.seed(999)
ahs_lasso_fs_1 = regularize_function(outcome = "LOGVALUE",
                        training = ahs_training,
                        testing =  ahs_testing,
                        FS_recipe = ahs_recipe_1,
                        type = "LASSO",
                        n_folds = 5,
                        grid_max = 30
                        )
ahs_lasso_fs_2 = regularize_function(outcome = "LOGVALUE",
                        training = ahs_training,
                        testing =  ahs_testing,
                        FS_recipe = ahs_recipe_2,
                        type = "LASSO",
                        n_folds = 10,
                        grid_max = 30
                        )

# Elastic Net -------------------- -------------------------------------
ahs_en_fs_1 = regularize_function(outcome = "LOGVALUE",
                        training = ahs_training,
                        testing =  ahs_testing,
                        FS_recipe = ahs_recipe_1,
                        type = "Elastic",
                        n_folds = 10,
                        grid_max = 30
                        )
ahs_en_fs_2 = regularize_function(outcome = "LOGVALUE",
                        training = ahs_training,
                        testing =  ahs_testing,
                        FS_recipe = ahs_recipe_2,
                        type = "Elastic",
                        n_folds = 10,
                        grid_max = 30
                        )
```

```{r}
# Assemble table 
ols_1 = ahs_ols_fs_1 %>% 
  mutate(Features = "Rep") %>%
  select(Model,Features,in_sample_rmse,oo_sample_rmse,is_r2,oos_r2,starts_with("P",ignore.case = F))
ols_2 = ahs_ols_fs_2 %>% 
  mutate(Features = "KS") %>%
  select(Model,Features,in_sample_rmse,oo_sample_rmse,is_r2,oos_r2,starts_with("P",ignore.case = F))
lasso_1 = ahs_lasso_fs_1 %>% 
  mutate(Features = "Rep") %>%
  select(Model,Features,in_sample_rmse,oo_sample_rmse,is_r2,oos_r2,starts_with("P",ignore.case = F))
lasso_2 = ahs_lasso_fs_2 %>% 
  mutate(Features = "KS") %>%
  select(Model,Features,in_sample_rmse,oo_sample_rmse,is_r2,oos_r2,starts_with("P",ignore.case = F))
elastic_1 = ahs_en_fs_1 %>% 
  mutate(Features = "Rep") %>%
  select(Model,Features,in_sample_rmse,oo_sample_rmse,is_r2,oos_r2,starts_with("P",ignore.case = F))
elastic_2 = ahs_en_fs_2 %>% 
  mutate(Features = "KS") %>%
  select(Model,Features,in_sample_rmse,oo_sample_rmse,is_r2,oos_r2,starts_with("P",ignore.case = F))

rbind(ols_1,
      ols_2,
      lasso_1,
      lasso_2,
      elastic_1, 
      elastic_2) %>% 
  mutate(
    # Calculate percent improvement
    P1 = ((ols_1$P1 - P1) / ols_1$P1) * 100,
    P2 = ((ols_1$P2 - P2) / ols_1$P2) * 100,
    P3 = ((ols_1$P3 - P3) / ols_1$P3) * 100,
    P4 = ((ols_1$P4 - P4) / ols_1$P4) * 100,
    P5 = ((ols_1$P5 - P5) / ols_1$P5) * 100,
  ) %>% 
  datasummary_df(
    align = "llccccccccc",
    title = "Replication of Mullinathan Spiess Table 1"
  )
```


# Question 3

```{r Prepare Binary Data}
training_g = G
training_g %<>% mutate(lbw = factor(ifelse(birth_weight <= 1500,"low","normal"),levels = c("low","normal")))
testing %<>%  mutate(lbw = factor(ifelse(birth_weight <= 1500,"low","normal"),  levels = c("low","normal")))
```

```{r Probit Small}

## N Y ------------------------------
tr = training_g %>% 
  group_by(lbw) %>% 
  summarise(n = n()) %>% 
  as_vector()
t = testing %>% 
  group_by(lbw) %>% 
  summarise(n = n()) %>% 
  as_vector()
is_n_y = tr[3]
os_n_y = t[3]
## Define Model ------------------------------
probit = logistic_reg() %>%
  set_engine("glm", 
             family = binomial(link = "probit")) 

## Define Recipe 
FS_recipe_1_lbw = recipe(as.formula(paste0("lbw","~",FS_1)), 
         data = training_g) %>%
         step_zv()
## Define Workflow ------------------------------

glm_workflow = 
 workflow() %>% 
  add_model(probit) %>% 
  add_recipe(FS_recipe_1_lbw)

## Fit Model -------------------------------------
fit_probit =
  glm_workflow %>% 
  fit(data = training_g)

## Obtain Predictions -----------------------------
probit_oos = predict(fit_probit,testing,type ="prob")
```

# b.

## Calibration Plot 

```{r Calibration Plot}
## Calibration Plot  ------------------------------
testing$y_hat_lbw = probit_oos$.pred_low
testing %<>% mutate(y_hat_b = ifelse(y_hat_lbw == "low",1,0))
# Get quantiles
q = quantile(testing$y_hat_lbw,probs = c(0,1))

testing %<>% mutate(low_bw = ifelse(lbw =="low",1,0))
  
calibration_1 = locpoly(x = testing$y_hat_lbw,
                        y = testing$low_bw, 
                        bandwidth = .3,
                        degree = 5,
                        kernel = "uniform",
                        range.x = c(q[1],q[2]),
                        gridsize = 1000)

calibration_1 %<>% 
  data.frame() 

ggplot() + 
  geom_line(size = 2, color = "blue",
            aes(calibration_1$x,calibration_1$y)) + 
  theme_bw() + 
  geom_abline(intercept = 0, 
              slope = 1, 
              size = 0.5,
              linetype = "dashed") + 
  labs(x = "Pr(BW = Low)",
       y = "Prop(Low)",
       title = "Calibration Plot (Small)")
```

The calinration plot displays the predicted probabilites of low-birthweight on the X axis and the share of observations classified as low-birthweight on the Y axis for each bin of predicted probabilites based on a local polynomial regression using a bandwidth of .3 and a uniform kernel. What we can broadly see is that at higher values of predicted low-birthweight, the model overpredicts the number of low-birthweight babies, while the true rates of low birthweight are lower than the 45 degree line. In the middle of the distribution of predicted probabilites, around .25-.6, the model tends to underpredict birthweight.

## Confusion Matrix 

```{r}
augment(fit_probit, new_data = testing) %>% 
  conf_mat(estimate = .pred_class,truth = lbw,)
```

For the confusion matrix, 67 true positives, 11420 true negatives, 51 false negatives, and 20 false positives. 

```{r}
## Accuracy
probit_oos_accuracy = augment(fit_probit, new_data = testing) %>% 
accuracy(truth = lbw, estimate = .pred_class)
# Calculate TPR or sensitivity 
probit_oos_sensitivity = augment(fit_probit, new_data = testing) %>% 
sens(truth = lbw, estimate = .pred_class)
## Precision 
probit_oos_precision = augment(fit_probit, new_data = testing) %>% 
precision(truth = lbw, estimate = .pred_class)
```

## ROC Curve

```{r}
 augment(fit_probit, new_data = testing) %>% 
  roc_curve(lbw,.pred_low) %>% 
  ggplot(aes(x = 1 - specificity, y = sensitivity)) +
  geom_path() +
  geom_abline(lty = 3) +
  coord_equal() +
  theme_bw() + 
  labs(title = "ROC Curve")

```
There is not a direct ROC curve interpretation. 

## AUC

```{r}
 auc = augment(fit_probit, new_data = testing) %>% 
  roc_auc(truth = lbw,.pred_low) 
```

```{r}
# Comine into table 
rbind(probit_oos_accuracy,
      probit_oos_precision,
      probit_oos_sensitivity,
      auc
      ) %>% 
  data.frame() %>% 
  datasummary_df(title = "Probit GOF (Feature set 1)")
```
The Probit GOF (Set 1) describes the results of various goodness of fit statistics. The first is the accuracy, which measures the share of correct classifications out of total classifications. The number correct is .99. Second is precision, which refers to the proportion of the models positive classifications that are actually positive, which has. value of .77.  The sensitivity is the true positive rate, which is the probability of a positive result given the case being positive, which is .57. Finally, I report the AUC which is .99. Relative to a model classifying at random, the model does quite well. 

## c. 

```{r Probit Large}

FS_recipe_3_lbw <- recipe(as.formula(paste0("lbw","~",FS_3)), 
                          data = training_g) %>%
  step_mutate(
    across(
      c(mom_race_hispanic_status, 
        mom_educ),
        as.factor
    )
  ) %>%
  step_novel(all_factor_predictors()) %>%
  step_dummy(all_nominal_predictors(), one_hot = TRUE) %>%
  step_zv(all_predictors())

## Define Workflow ------------------------------
model_tune_probit <- logistic_reg(
  penalty = tune(),
  mixture = 1
) %>%
  set_engine("glmnet", family = "binomial", link = "probit") %>%
  set_mode("classification")

# Cross-Validate ------------------------------

folds <- vfold_cv(training_g, v = 5)
lambda_grid <- expand.grid(penalty = c(0, 100))

cv_lambda <- tune_grid(
  model_tune_probit,
  FS_recipe_3_lbw,
  resamples = folds,
  grid = lambda_grid
)

best <- select_best(cv_lambda, metric = "roc_auc")[1,1]

final_wf <- workflow() %>%
  add_recipe(FS_recipe_3_lbw) %>%
  add_model(model_tune_probit)

final_fit <- finalize_workflow(final_wf, best) %>%
  fit(training_g)

## Obtain Predictions -----------------------------
probit_oos_lg = predict(final_fit,testing,type ="prob")
## Calibration plot -----------------------------

testing$y_hat_lbw = probit_oos_lg$.pred_low
testing %<>% mutate(y_hat_b = ifelse(y_hat_lbw == "low",1,0))
# Get quantiles
q = quantile(testing$y_hat_lbw,probs = c(0,1),na.rm =  T)
testing %<>% mutate(low_bw = ifelse(lbw =="low",1,0))
  
calibration_1 = locpoly(x = testing$y_hat_lbw,
                        y = testing$low_bw, 
                        bandwidth = .3,
                        degree = 5,
                        kernel = "uniform",
                        range.x = c(q[1],q[2]),
                        gridsize = 1000)

calibration_1 %<>% 
  data.frame() 

ggplot() + 
  geom_line(size = 2, color = "blue",
            aes(calibration_1$x,calibration_1$y)) + 
  theme_bw() + 
  geom_abline(intercept = 0, 
              slope = 1, 
              size = 0.5,
              linetype = "dashed") + 
  labs(x = "Pr(BW = Low)",
       y = "Prop(Low)",
       title = "Calibration Plot (Large Feature Set)")
## Goodness of Fit  -----------------------------
### Confusion Matrix 
augment(final_fit, new_data = testing) %>%
  conf_mat(truth = lbw, estimate = .pred_class)
### Accuracy
probit_oos_accuracy_lg = augment(final_fit, new_data = testing) %>% 
accuracy(truth = lbw, estimate = .pred_class)
### Calculate TPR or sensitivity 
probit_oos_sensitivity_lg = augment(final_fit, new_data = testing) %>% 
sens(truth = lbw, estimate = .pred_class)
### Precision 
probit_oos_precision_lg = augment(final_fit, new_data = testing) %>% 
precision(truth = lbw, estimate = .pred_class)
### AUC
 auc_lg = augment(final_fit, new_data = testing) %>% 
  roc_auc(truth = lbw,.pred_low) 
### ROC Curve 
 augment(final_fit, new_data = testing) %>% 
  roc_curve(lbw,.pred_low) %>% 
  ggplot(aes(x = 1 - specificity, y = sensitivity)) +
  geom_path() +
  geom_abline(lty = 3) +
  coord_equal() +
  theme_bw() + 
  labs(title = "ROC Curve")
 
### Display GOF 
 rbind(probit_oos_accuracy_lg,
       probit_oos_precision_lg,
       probit_oos_sensitivity_lg,
       auc_lg
      ) %>% 
  data.frame() %>% 
  datasummary_df(title = "Probit GOF (Feature set 3)")

```

For the confusion matrix, 64 true positives, 11397 true negatives, 54 false negatives, and 23 false positives. 

The Probit GOF (Set 3) table describes the results of various goodness of fit statistics. The first is the accuracy, which measures the share of correct classifications out of total classifications. The number correct is .99. Second is precision, which refers to the proportion of the models positive classifications that are actually positive, which has. value of .74.  The sensitivity is the true positive rate, which is the probability of a positive result given the case being positive, which is .54. Finally, I report the AUC which is .98. Relative to the small feature set, this model does a little bit worse at the true positive rate (sensitivity) and the AUC. But it is not dramatic.






