---
title: "PUBPOL 5390 Homework 2"
author: Thelonious Goerz
date: "Draft date: `r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document:
    number_sections: yes
linestretch: 1.5
fig_width: 5
fig_height: 2.5
csl: "/Users/theloniousgoerz/Academic/Professional/Resources/Citations/styles-master/american-sociological-association.csl"
editor_options:
  chunk_output_type: inline
geometry: "left=1in,right=1in,top=1in,bottom=1in"
number-sections: true
keep_tex: true
link-citations: yes
linkcolor: blue
font_size: 12pt
header-includes: 
  - \usepackage{framed} 
  - \usepackage{amsmath}
  - \usepackage{tcolorbox} 
  - \usepackage{pdflscape}
  - \newcommand{\blandscape}{\begin{landscape}}
  - \newcommand{\elandscape}{\end{landscape}}
  - \usepackage{multirow} 
  - \usepackage{xcolor} 
  - \usepackage{tabularray}
  - \def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}} 
  - \usepackage{setspace} 
  - \usepackage{threeparttable}
  - \usepackage{booktabs}
  - \usepackage{siunitx}
  - \usepackage{tikz}
  - \usepackage{tabularray}
  - \usepackage{float}
  - \usepackage{graphicx}
  - \usepackage{rotating}
  - \usepackage[normalem]{ulem}
  - \UseTblrLibrary{booktabs}
  - \UseTblrLibrary{siunitx}
---

```{r setup, include=TRUE, results="hide", echo = FALSE, message = F}

 rm(list = objects())
 # THIS IS THE WIDTH OF THE CODE                                                ###
 knitr::opts_chunk$set(echo = F,
                       resuls = T, message = F, warning = F)
 
 # knitr::opts_chunk$set(fig.width = unit(16,"cm"), fig.height = unit(9,"cm"))
 
```

\doublespacing
\fontsize{12}{22}
\setstretch{2}
\baselineskip=29pt

```{r Load Packages}
rm(list = ls())
### Packages ###
library(readr)
library(tidyverse)
library(fixest)
library(modelsummary)
library(haven)
library(magrittr)
library(splines)
library(glmnet)
library(KernSmooth)
library(boot)
library(tidymodels)
library(here)
library(KernSmooth)

```

```{r Data}
# Problem 1 
Gsmall = read_csv(here("Data","births_for_class_Gsmall.csv"))
H = read_csv(here("Data","births_for_class_H.csv"))
G = read_csv(here("Data","births_for_class_G.csv"))
```

# Question 1

```{r Feature Set}
set1 = c(
    "gestation_weeks",
    "mom_age",
    "mom_race_hispanic_status",
    "mom_educ")

set2 = c(    
    "mom_age",
    "mom_race_hispanic_status",
    "mom_educ",
    "gestation_weeks",
    "time",
    "in_hospital",
    "gestation_weeks"
    )

set3 = c(    
    "mom_age",
    "mom_race_hispanic_status",
    "mom_educ",
    "gestation_weeks",
    "time",
    "in_hospital",
    "gestation_weeks",
    "prior_births_living",
    "plurality",
    "residence_status",
    "mom_bmi",
    "dad_age"
    )

FS_1 = paste(
set1,
collapse = "+"
)

FS_2 = paste(
set2,
collapse = "+"
)

FS_3 = paste(
set3,
collapse = "+"
)
```

```{r}
# Label Train and Test Set 
training = Gsmall
testing = H
```

```{r OLS}
ols_function = function(outcome,
                        training,
                        testing,
                        FS_recipe
                        ){
## SD Y ------------------------------
is_sd_y = sd(training[[outcome]])
os_sd_y = sd(testing[[outcome]])
## Define Model ------------------------------
ols = linear_reg(
  mode = "regression",
  engine = "lm"
) 
## Define Workflow ------------------------------
ols_workflow = 
 workflow() %>% 
  add_model(ols) %>% 
  add_recipe(FS_recipe)

## Fit Model -------------------------------------
fit_ols =
  ols_workflow %>% 
  fit(data = training)

## Obtain Predictions -----------------------------
ols_oos = predict(fit_ols,testing)
  
# Calculate RMSE ---------------------------------

## Combine Train-Test predictions and Outcomes into a DF 
ols_gof_train_df = data.frame(
  estimate = as.numeric(predict(fit_ols,training)$.pred),
  truth = unlist(fit_ols$pre$mold$outcomes[[outcome]]))

ols_gof_test_df = data.frame(
  estimate = as.numeric(ols_oos$.pred),
  truth = as.numeric(testing[[outcome]]))

## Calculate RMSE ------------------------------
ols_train_rmse = rmse(
     data = ols_gof_train_df,
     truth,
     estimate)

ols_test_rmse = rmse(
     data = ols_gof_test_df,
     truth,
     estimate)
# Return Items ------------------------------
return(
  data.frame(
    Model = "OLS",
    in_sample_sd_y = is_sd_y,
    oo_sample_sd_y = os_sd_y,
    in_sample_rmse = ols_train_rmse$.estimate,
    oo_sample_rmse = ols_test_rmse$.estimate,
    n_predictors = ncol(fit_ols$pre$mold$predictors)
  )
)
}
```

```{r Run OLS For Each Feature Set}
## Create Recipes 
FS_recipe_1 = recipe(as.formula(paste0("birth_weight","~",FS_1)), 
         data = training) %>%
         step_zv()

FS_recipe_2 = recipe(as.formula(paste0("birth_weight","~",FS_2)), 
         data = training) %>%
  step_mutate(
    across(c(
    mom_race_hispanic_status,
    mom_educ,
    gestation_weeks
    ), 
    as.factor)
  ) %>% 
    step_dummy(all_nominal_predictors(),
             one_hot = T) %>% 
  step_zv(all_predictors())

FS_recipe_3 = recipe(as.formula(paste0("birth_weight","~",FS_3)), 
         data = training) %>%
  step_mutate(
    across(c(
    mom_race_hispanic_status,
    mom_educ,
    gestation_weeks,
    time,
    ), 
    as.factor)
  ) %>% 
    step_dummy(all_nominal_predictors(),
                one_hot = T) %>% 
  step_spline_b(mom_age,
                deg_free = 6,
                degree = 3,
                keep_original_cols = T) %>%
  step_zv(all_predictors())

# Run OLS Function on feature sets 
ols_fs_1 = ols_function("birth_weight",
    FS_recipe = FS_recipe_1,
    training = training,
    testing = testing)

ols_fs_2 = ols_function("birth_weight",
    FS_recipe = FS_recipe_2,
    training = training,
    testing = testing) 

ols_fs_3 = ols_function("birth_weight",
    FS_recipe = FS_recipe_3,
    training = training,
    testing = testing) 

### Save Metrics ###

rbind(ols_fs_1,
      ols_fs_2,
      ols_fs_3)
# why are these not chnaging? hmm 
```

```{r LASSO}
regularize_function = function(outcome,
                        training,
                        testing,
                        FS_recipe,
                        type,
                        n_folds,
                        grid_max
                        ){
## SD Y ------------------------------
is_sd_y = sd(training[[outcome]])
os_sd_y = sd(testing[[outcome]])
## Tune Parameters ------------------------------
## Pick Model

## Define folds 
folds = vfold_cv(training,v = n_folds) 

if (type == "LASSO"){
model_tune = linear_reg(
  mode = "regression",
  engine = "glmnet",
  mixture = 1, 
  penalty = tune()
) 
# Cross-Validate
cv_lambda = tune_grid(model_tune, 
                      FS_recipe,
                      resamples = folds, 
                      tune = penalty,
                      grid = expand.grid(penalty = 0:grid_max))

# Pick lambda # 
best_penalty = as.numeric(show_best(cv_lambda)[1,"penalty"])
best_mixture = 1

} else if (type == "Ridge"){
model_tune = linear_reg(
  mode = "regression",
  engine = "glmnet",
  mixture = 0, 
  penalty = tune()
) 
# Cross-Validate
cv_lambda = tune_grid(model_tune, 
                      FS_recipe,
                      resamples = folds, 
                      tune = penalty,
                      grid = expand.grid(penalty = 0:grid_max))

# Pick lambda # 
best_penalty = as.numeric(show_best(cv_lambda)[1,"penalty"])
best_mixture = 0

  
} else if (type == "Elastic"){
model_tune = linear_reg(
  mode = "regression",
  engine = "glmnet",
  mixture = tune(), 
  penalty = tune()
) 
# Cross-Validate 

cv_pick = tune_grid(model_tune, 
                      FS_recipe,
                      resamples = folds, 
                      tune = c(penalty,mixture),
                      grid = expand.grid(penalty = 0:grid_max,
                                         mixture = c(.25,.5,.75)))

# Pick lambda # 
best_penalty = as.numeric(show_best(cv_pick)[1,"penalty"])
best_mixture = as.numeric(show_best(cv_pick)[1,"mixture"])
}

# ------------------------------------------------------------
# Declare Model # 
regularized_model = linear_reg(
  mode = "regression",
  engine = "glmnet",
  mixture = best_mixture, 
  penalty = best_penalty
) 
## Define Workflow ------------------------------
rglr_workflow = 
 workflow() %>% 
  add_model(regularized_model) %>% 
  add_recipe(FS_recipe)

## Fit Model -------------------------------------
fit_rglr =
  rglr_workflow %>% 
  fit(data = training)

## Obtain Predictions -----------------------------
rglr_oos = predict(fit_rglr,testing)

tidy_coefficients <- fit_rglr %>%
      extract_fit_parsnip() %>%
      tidy()
    # Filter for non-zero coefficients (selected predictors)
    n_terms <- tidy_coefficients %>%
      filter(estimate != 0) %>%
      pull(term) %>% 
      length() %>% 
      as.numeric()
  
# Calculate RMSE ---------------------------------
## Combine Train-Test predictions and Outcomes into a DF 
rglr_gof_train_df = data.frame(
  estimate = as.numeric(predict(fit_rglr,training)$.pred),
  truth = unlist(fit_rglr$pre$mold$outcomes[[outcome]]))

rglr_gof_test_df = data.frame(
  estimate = as.numeric(rglr_oos$.pred),
  truth = as.numeric(testing[[outcome]]))

## Calculate RMSE ------------------------------
rglr_train_rmse = rmse(
     data = rglr_gof_train_df,
     truth,
     estimate)

rglr_test_rmse = rmse(
     data = rglr_gof_test_df,
     truth,
     estimate)
# Return Items ------------------------------
 return(
   data.frame(
     Model = {{type}},
     in_sample_sd_y = is_sd_y,
     oo_sample_sd_y = os_sd_y,
     in_sample_rmse = rglr_train_rmse$.estimate,
     oo_sample_rmse = rglr_test_rmse$.estimate,
     n_predictors = n_terms,
     mixture = best_mixture,
     penalty = best_penalty
   )
 ) 

}

```

```{r Run Regularized Regressions}
set.seed(9999)
# Run LASSO Models ------------------------------------- # 
# 
# -------------------------------------------------------#

lasso_FS_1 = regularize_function(outcome = "birth_weight",
                        training = training,
                        testing =  testing,
                        FS_recipe = FS_recipe_1,
                        type = "LASSO",
                        n_folds = 10,
                        grid_max = 30
                        )

lasso_FS_2 = regularize_function(outcome = "birth_weight",
                        training = training,
                        testing =  testing,
                        FS_recipe = FS_recipe_2,
                        type = "LASSO",
                        n_folds = 10,
                        grid_max = 30
                        )

lasso_FS_3 = regularize_function(outcome = "birth_weight",
                        training = training,
                        testing =  testing,
                        FS_recipe = FS_recipe_3,
                        type = "LASSO",
                        n_folds = 10,
                        grid_max = 30
                        )

set.seed(9999)
# Run RIDGE Models ------------------------------------- # 
# 
# -------------------------------------------------------#

ridge_FS_1 = regularize_function(outcome = "birth_weight",
                        training = training,
                        testing =  testing,
                        FS_recipe = FS_recipe_1,
                        type = "Ridge",
                        n_folds = 10,
                        grid_max = 50
                        )

ridge_FS_2 = regularize_function(outcome = "birth_weight",
                        training = training,
                        testing =  testing,
                        FS_recipe = FS_recipe_2,
                        type = "Ridge",
                        n_folds = 10,
                        grid_max = 50
                        )

ridge_FS_3 = regularize_function(outcome = "birth_weight",
                        training = training,
                        testing =  testing,
                        FS_recipe = FS_recipe_3,
                        type = "Ridge",
                        n_folds = 10,
                        grid_max = 50
                        )

# Run Elastic Net Models ------------------------------------- # 
# 
# -------------------------------------------------------------#

elastic_FS_1 = regularize_function(outcome = "birth_weight",
                        training = training,
                        testing =  testing,
                        FS_recipe = FS_recipe_1,
                        type = "Elastic",
                        n_folds = 10,
                        grid_max = 50
                        )

elastic_FS_2 = regularize_function(outcome = "birth_weight",
                        training = training,
                        testing =  testing,
                        FS_recipe = FS_recipe_2,
                        type = "Elastic",
                        n_folds = 10,
                        grid_max = 50
                        )

elastic_FS_3 = regularize_function(outcome = "birth_weight",
                        training = training,
                        testing =  testing,
                        FS_recipe = FS_recipe_3,
                        type = "Elastic",
                        n_folds = 10,
                        grid_max = 50
                        )


ols_fs_1 %<>% mutate(mixture = NA,penalty = NA)
ols_fs_2 %<>% mutate(mixture = NA,penalty = NA)
ols_fs_3 %<>% mutate(mixture = NA,penalty = NA)

# Look at Models

rbind(ols_fs_1,
      ols_fs_2,
      ols_fs_3,
      lasso_FS_1,
      lasso_FS_2,
      lasso_FS_3,
      # Ridge
      ridge_FS_1,
      ridge_FS_2,
      ridge_FS_3,
      # Elastic
      elastic_FS_1,
      elastic_FS_2,
      elastic_FS_3
      ) %>% 
  datasummary_df(
    title = "Comparison of Predictive Performance Across Models and Feature Sets"
  )
```
### Cross-Validation 

Within the tidymodels framework (the R analogue to scikit-learn), there are no defaults for cross-validation. Instead, I wrote my own function a la Mullinathan and Spiess, to run the models in an automated way. These functions allow the grid size for parameters to be searched over and the number of folds for cross-validation. For all models, I set the number of folds to 10, the grid size to 30 or 50 depending on whether the penalty is L1 or L2. The goodness of fit measure that I use is RMSE. 

### Model Performance 

I use three different covariate sets that layer controls in each model and change the functional form. The firs model includes linear forms of relevant predictors (e.g., mother's age and gestation). The second and third incorporate non-linear terms as well as expanded covariates. The majority of the improvement in predictive performance in richer models is obtained from gestation weeks. 

Overall, the results indicate that the best model in terms of predictive performance is a LASSO regression that selects 30 predictors out of a candidate set of 39 and achieves an out-of-sample RMSE of $437$. Across ridge and elastic net, the RMSE is slightly worse but not by that much. In general, each of the second feature set models achieves similar results. The third feature set, which is the richest model, performs the worst across models. Though, the LASSO model shrinks the feature set from 542 to 94, which reduces the RMSE. Even with shrinkage, it still performs poorly. 

# Question 2

```{r}

```

# Question 3

```{r Prepare Binary Data}
training_g = G
training_g %<>% mutate(lbw = factor(ifelse(birth_weight <= 1500,"low","normal"),levels = c("low","normal")))
testing %<>%  mutate(lbw = factor(ifelse(birth_weight <= 1500,"low","normal"),  levels = c("low","normal")))
```

```{r Probit Small}

## N Y ------------------------------
tr = training_g %>% 
  group_by(lbw) %>% 
  summarise(n = n()) %>% 
  as_vector()
t = testing %>% 
  group_by(lbw) %>% 
  summarise(n = n()) %>% 
  as_vector()
is_n_y = tr[3]
os_n_y = t[3]
## Define Model ------------------------------
probit = logistic_reg() %>%
  set_engine("glm", 
             family = binomial(link = "probit")) 

## Define Recipe 
FS_recipe_1_lbw = recipe(as.formula(paste0("lbw","~",FS_1)), 
         data = training_g) %>%
         step_zv()
## Define Workflow ------------------------------

glm_workflow = 
 workflow() %>% 
  add_model(probit) %>% 
  add_recipe(FS_recipe_1_lbw)

## Fit Model -------------------------------------
fit_probit =
  glm_workflow %>% 
  fit(data = training_g)

## Obtain Predictions -----------------------------
probit_oos = predict(fit_probit,testing,type ="prob")
```

# b.

## Calibration Plot 

```{r Calibration Plot}
## Calibration Plot  ------------------------------
testing$y_hat_lbw = probit_oos$.pred_low
testing %<>% mutate(y_hat_b = ifelse(y_hat_lbw == "low",1,0))
# Get quantiles
q = quantile(testing$y_hat_lbw,probs = c(0,1))

testing %<>% mutate(low_bw = ifelse(lbw =="low",1,0))
  
calibration_1 = locpoly(x = testing$y_hat_lbw,
                        y = testing$low_bw, 
                        bandwidth = .3,
                        degree = 5,
                        kernel = "uniform",
                        range.x = c(q[1],q[2]),
                        gridsize = 1000)

calibration_1 %<>% 
  data.frame() 

ggplot() + 
  geom_line(size = 2, color = "blue",
            aes(calibration_1$x,calibration_1$y)) + 
  theme_bw() + 
  geom_abline(intercept = 0, 
              slope = 1, 
              size = 0.5,
              linetype = "dashed") + 
  labs(x = "Pr(BW = Low)",
       y = "Prop(Low)",
       title = "Calibration Plot")
```

## Confusion Matrix 

```{r}
augment(fit_probit, new_data = testing) %>% 
  conf_mat(truth = lbw, estimate = .pred_class) 

 augment(fit_probit, new_data = testing) %>% 
   group_by(.pred_class,lbw) %>% 
   summarise(n())
```

```{r}
## Accuracy
probit_oos_accuracy = augment(fit_probit, new_data = testing) %>% 
accuracy(truth = lbw, estimate = .pred_class)
# Calculate TPR or sensitivity 
probit_oos_sensitivity = augment(fit_probit, new_data = testing) %>% 
sens(truth = lbw, estimate = .pred_class)
## Precision 
probit_oos_precision = augment(fit_probit, new_data = testing) %>% 
precision(truth = lbw, estimate = .pred_class)
```

## ROC Curve

```{r}
 augment(fit_probit, new_data = testing) %>% 
  roc_curve(lbw,.pred_low) %>% 
  ggplot(aes(x = 1 - specificity, y = sensitivity)) +
  geom_path() +
  geom_abline(lty = 3) +
  coord_equal() +
  theme_bw() + 
  labs(title = "ROC Curve")

```
## AUC

```{r}
 auc = augment(fit_probit, new_data = testing) %>% 
  roc_auc(truth = lbw,.pred_low) 
```

```{r}
# Comine into table 
rbind(probit_oos_accuracy,
      probit_oos_precision,
      probit_oos_sensitivity,
      auc
      ) %>% 
  data.frame() %>% 
  datasummary_df(title = "Probit GOF (Feature set 1)")
```

## c. 

```{r Probit Large}

FS_recipe_3_lbw <- recipe(as.formula(paste0("lbw","~",FS_3)), 
                          data = training_g) %>%
  step_mutate(
    across(
      c(mom_race_hispanic_status, 
        mom_educ),
        as.factor
    )
  ) %>%
  step_novel(all_factor_predictors()) %>%
  step_dummy(all_nominal_predictors(), one_hot = TRUE) %>%
  step_zv(all_predictors())

## Define Workflow ------------------------------
model_tune_probit <- logistic_reg(
  penalty = tune(),
  mixture = 1
) %>%
  set_engine("glmnet", family = "binomial", link = "probit") %>%
  set_mode("classification")

# Cross-Validate ------------------------------

folds <- vfold_cv(training_g, v = 5)
lambda_grid <- expand.grid(penalty = c(0, 100))

cv_lambda <- tune_grid(
  model_tune_probit,
  FS_recipe_3_lbw,
  resamples = folds,
  grid = lambda_grid
)

best <- select_best(cv_lambda, metric = "roc_auc")[1,1]

final_wf <- workflow() %>%
  add_recipe(FS_recipe_3_lbw) %>%
  add_model(model_tune_probit)

final_fit <- finalize_workflow(final_wf, best) %>%
  fit(training_g)

## Obtain Predictions -----------------------------
probit_oos_lg = predict(fit_probit_lg,testing,type ="prob")
## Calibration plot -----------------------------

testing$y_hat_lbw = probit_oos_lg$.pred_low
testing %<>% mutate(y_hat_b = ifelse(y_hat_lbw == "low",1,0))
# Get quantiles
q = quantile(testing$y_hat_lbw,probs = c(0,1),na.rm =  T)
testing %<>% mutate(low_bw = ifelse(lbw =="low",1,0))
  
calibration_1 = locpoly(x = testing$y_hat_lbw,
                        y = testing$low_bw, 
                        bandwidth = .3,
                        degree = 5,
                        kernel = "uniform",
                        range.x = c(q[1],q[2]),
                        gridsize = 1000)

calibration_1 %<>% 
  data.frame() 

ggplot() + 
  geom_line(size = 2, color = "blue",
            aes(calibration_1$x,calibration_1$y)) + 
  theme_bw() + 
  geom_abline(intercept = 0, 
              slope = 1, 
              size = 0.5,
              linetype = "dashed") + 
  labs(x = "Pr(BW = Low)",
       y = "Prop(Low)",
       title = "Calibration Plot (Large Feature Set)")
## Goodness of Fit  -----------------------------
### Confusion Matrix 
augment(final_fit, new_data = testing) %>%
  conf_mat(truth = lbw, estimate = .pred_class)
### Accuracy
probit_oos_accuracy_lg = augment(final_fit, new_data = testing) %>% 
accuracy(truth = lbw, estimate = .pred_class)
### Calculate TPR or sensitivity 
probit_oos_sensitivity_lg = augment(final_fit, new_data = testing) %>% 
sens(truth = lbw, estimate = .pred_class)
### Precision 
probit_oos_precision_lg = augment(final_fit, new_data = testing) %>% 
precision(truth = lbw, estimate = .pred_class)
### AUC
 auc_lg = augment(final_fit, new_data = testing) %>% 
  roc_auc(truth = lbw,.pred_low) 
### ROC Curve 
 augment(final_fit, new_data = testing) %>% 
  roc_curve(lbw,.pred_low) %>% 
  ggplot(aes(x = 1 - specificity, y = sensitivity)) +
  geom_path() +
  geom_abline(lty = 3) +
  coord_equal() +
  theme_bw() + 
  labs(title = "ROC Curve")
 
### Display GOF 
 rbind(probit_oos_accuracy_lg,
       probit_oos_precision_lg,
       probit_oos_sensitivity_lg,
       auc_lg
      ) %>% 
  data.frame() %>% 
  datasummary_df(title = "Probit GOF (Feature set 3)")

```








