---
title: "Self-Fullfilling Prophecy?: Do Subjective Mortality Expectations Matter For Predictive Mortality and Longevity"
author: |
  | Thelonious Goerz
  | PUBPOL 5390
date: "Draft date: `r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document:
    number_sections: yes
linestretch: 1.5
fig_width: 5
fig_height: 2.5
csl: "/Users/theloniousgoerz/Academic/Professional/Resources/Citations/styles-master/american-sociological-association.csl"
editor_options:
  chunk_output_type: inline
  markdown: 
    wrap: 72
geometry: "left=1in,right=1in,top=1in,bottom=1in"
number-sections: true
keep_tex: true
link-citations: yes
linkcolor: blue
font_size: 12pt
header-includes: 
  - \usepackage{framed} 
  - \usepackage{amsmath}
  - \usepackage{tcolorbox} 
  - \usepackage{pdflscape}
  - \newcommand{\blandscape}{\begin{landscape}}
  - \newcommand{\elandscape}{\end{landscape}}
  - \usepackage{multirow} 
  - \usepackage{xcolor} 
  - \usepackage{tabularray}
  - \def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}} 
  - \usepackage{setspace} 
  - \usepackage{threeparttable}
  - \usepackage{booktabs}
  - \usepackage{siunitx}
  - \usepackage{tikz}
  - \usepackage{tabularray}
  - \usepackage{float}
  - \usepackage{graphicx}
  - \usepackage{rotating}
  - \usepackage[normalem]{ulem}
  - \UseTblrLibrary{booktabs}
  - \UseTblrLibrary{siunitx}
  - \newcommand{\tinytableTabularrayUnderline}[1]{\underline{#1}}
  - \newcommand{\tinytableTabularrayStrikeout}[1]{\sout{#1}}
  - \NewTableCommand{\tinytableDefineColor}[3]{\definecolor{#1}{#2}{#3}}
bibliography: references.bib
---

```{r setup, include=TRUE, results="hide", echo = FALSE, message = F}

 rm(list = objects())
 # THIS IS THE WIDTH OF THE CODE                                                ###
 knitr::opts_chunk$set(echo = F,
                       resuls = T, message = F, warning = F)
 
 # knitr::opts_chunk$set(fig.width = unit(16,"cm"), fig.height = unit(9,"cm"))
 library(tidyverse)
 library(modelsummary)
 library(kableExtra)
 
```

\textbf{Abstract}

Subjective expectations about health are thought to matter theoretically
for mortality outcomes because they contain rich, but unobserved
information about health status. While subjective expectations are
correlated with mortality outcomes, whether they enrich the prediction
of individual longevity remains less clear. This paper examines whether
subjective expectations improve the predictability of longevity compared
to sociodemographic and economic covariates using representative panel
data and a variety of supervised learning models. Results indicate that
subjective expectations contribute little predictive power to models
predicting death age, but improve predictions of survival past the age
of 75. These results suggest that health surveys may be enriched by
incorporating subjective mortality questions.

\doublespacing

\section{Introduction}

```{r, echo = F, results = F}
rmdwc::rmdcountAddin()
```

The extent to which we can predict mortality well matters for
understanding population health and policy decisions to meet the growth
of the aging population in the U.S. Recent demographic work has thus
aimed to address this problem by leveraging various demographic,
forecasting, and prediction methods to study the predictability of
longevity. However, recent work shows that individual-level mortality
and longevity prediction is challenging, with uncertainty remaining
high, even with big microdata and rich individual-level covariates
[@breen_demographic_2023]. Moreover, even in cases where predictive
performance is better (i.e., with survival curves), there are notable
inequalities in predictive performance across group [@badolato2023].

Despite empirical challenges, obtaining better predictions of
individual-mortality and longevity are crucial for policy and scholarly
reasons. First, predicting mortality from individual-level data is
important for forecasting population longevity in the future, especially
by social groups or categories that are not readily available from
tabular data. Generating accurate predictions can inform public health
policy decisions allocating resources to populations ar older ages based
on predicted life expectancy with morbidities, disabilities, disease,
loneliness, and other chronic conditions [@raymo2022]. Second, because
research documents notable inequalities in the predictability of
mortality for different social groups, the incorporation of subjective
measures may improve predictive accuracy and thereby contribute to
efforts to narrow mortality and longevity inequalities (e.g., by race).

One potential avenue for improving predictions, is to leverage datasets,
such as the Health and Retirement Study (HRS), that contain rich
covariates as well as objective (e.g., functional limitations) and
subjective health information. Past work has employed subjective
mortality expectation questions in the HRS to aid in interpreting the
predictive performance of models, but has overlooked their role as focal
features. While current evidence suggests that individuals are not very
good at predicting their own mortality and there are noted inequalities
in performance by social group, whether subjective probabilities can be
used in a prediction model as "utility features" [@kose] in combination
with strongly predictive variables (e.g., education, income, age,
health) to achieve gains in predictive performance is currently
unexplored.

The present study builds upon past efforts by testing the predictive
performance of subjective mortality measures to (1) predict individual
longevity (death age in years) and (2), as a secondary exercise, predict
survival past threshold ages (e.g., 75 years) relative to models without
expectations. To examine the predictability of mortality and longevity,
I leverage HRS panel data and a variety of machine learning models to
evaluate predictive performance relative to a baseline OLS regression
model, predicting longevity and living to 75 years as a function of
subjective probabilities only. For each model, I contrast two feature
sets: the first is a set of 82 cross-sectional and panel covariates, and
the second feature set adds the subjective information to the 82
covariates.[^1]

[^1]: Note, the 82 covariates refer to variables in their "unprocessed"
    form. That is, before creating factors, deciles, and nonlinear
    transformations of the variables.

Results from random forest and XGBoost models combined with rich, panel
data reveal that subjective measures do little to improve the
predictability of longevity over other demographic, economic, and health
covariates. However, subjective expectations do improve the
predictability of surviving to age 75 compared to covariates.
Comparisons in the specific predictive performance algorithms and
variable importance are discussed.

\section{Background}

The population in the United States and across the world is aging
rapidly as human lifespans continue to extend well beyond previously
thought limits. A recent report by the Pew Research Center
[-@schaeffer2024] projected that the population of adults 65 years and
older will quadruple by 2054. An increasingly old population poses many
problems related to caring for the old and ensuring enough labor supply
to sustain economic productivity. Accordingly, questions related to how
long we will live and what factors predict longer and shorter lifespans
have become increasingly salient to policymakers, public health
researchers, and demographers. An emergent literature in demography has
reformulated these questions as a prediction problems that can tackled
with sophisticated machine learning methods that leverage advancements
in computational capacity [@breen_demographic_2023; @badolato2023].

Using a variety of datasets ranging from the full count census to the
Health and Retirement Study (HRS), evidence demonstrates that predicting
longevity is challenging, even with large datasets and rich predictors.
Despite a wealth of social science evidence on the determinants of
mortality, obtaining good predictions across diverse populations is
challenging. Breen and Seltzer [-@breen_demographic_2023] show that in
both cohort and period analyses predicting the probability that an
individual will die in the next ten years, the $R^3$ for preferred
prediction models is low, with as much as $98.6\%$ of the variance
unexplained. Using the HRS, @badolato2023 found better predictive
improvements when modeling Cox survival curves, which partly arises from
the rich covariate set and longitudinal nature of the data. However,
they document stark inequalities in predictability by social group,
particularly for race, gender, and education.

A common takeaway of both papers is that mortality and longevity can be
unpredictable and prediction quality varies. First, Sociodemographic
features provide weak predictive information, despite substantive and
theoretical importance [@breen_demographic_2023]. Second, researchers
need to employ a more expansive set of theoretical predictors guided by
theory and practical intuition. This paper attempts to make headway on
improving our ability to predict mortality and longevity by considering
the potential value of individuals subjective beliefs about their own
mortality. While @breen_demographic_2023 and @badolato2023 discuss
subjective mortality information, they do so in the context of
benchmarking estimates but do not explicitly evaluate their predictive
performance.

There are a variety of theoretical and empirical reasons why subjective
information may be valuable for predicting longevity. First, decades of
sociological and economic research recognizes the theoretically
predictive power of individual beliefs on actual outcomes. The concept
of the self-fulfilling prophecy, originally coined by @merton1948, which
states that beliefs about future events, whether true or false, can
engender those events. In the context of health and longevity, this can
be seen intuitively by considering that individuals generally have a
good sense of their own health and have knowledge about their own
individual circumstances that are unobserved by covariates but may
predict mortality. Second, in the context of predicting individual
longevity, subjective information about health is relatively low cost to
collect. Rather than asking complex health screening questions or
collecting expensive biomarkers, asking a simple set of questions about
subjective views of mortality is relatively easy and low-cost to
implement on health surveys. This is one reason why self-rated health
questions are common on surveys: they have been validated in various
settings and across groups and are predictive of future mortality.

In research on aging and health, subjective mortality probabilities have
been employed before in forecasting settings, yet a full assessment of
their predictive utility remains far from settled. For instance, one
study previously used subjective measures in the HRS to calculate
subjective, cohort-life-table estimates for U.S. men and women that were
found to be predictive of revisions to Social Security life table
estimates [@perozek2008]. Indeed, evidence suggests that subjective
expectations have the potential to contribute meaningful information,
yet their contributions to predictive exercises remain unclear.

\section{Data}

\subsection{Data Description}

To predict longevity, I leverage the Health and Retirement Study (HRS) –
a panel dataset of approximately 32,000 U.S. adults 50 years and older.
The HRS contains comprehensive information on health, wealth, stock
portfolios, social and personal relationships, and other variables
related to aging. The HRS also tracks mortality outcomes of sample
participants that form the basis of this study. For this exercise, I
focus on the sample of individuals who have died and collapse the long
panel dataset into a wide dataset, where each row indexes an individual
with a single death age and columns index both time invariant covariates
(e.g., race, education) and time varying covariates (e.g., income at
each wave). This transformation is also advantageous for bootstrapping
and cross-validation, because it ensures that individuals' entire sets
of person-observations are sampled into a fold or sample to avoid
sampling individuals at different time points into different folds or
samples, which could pose issues for prediction. Table 1 describes a set
of focal (but not all) covariates in detail before pre-processing.

\input{./Output/01_descriptive_table.tex}

In addition to sociodemographic, economic, and health features, the HRS
includes sets of subjective expectations about retirement planning,
financial planning, and health. For mortality, subjective expectation
questions ask respondents to rate the probability from 0-100 that they
will live beyond a certain age (i.e., 75 and 80 years of age)
[@perozek2008; @hurd2002]. Specifically, the question asks: "What do you
think are the chances that you will live to be 75 or more?" Using the
subjective indicators, I test whether they have predictive power for
both longevity and survival past 75 years of age. To do so, I examine
how well machine learning models leveraging all non-subjective features
(Covariate set 1) perform compared to models that include subjective
features (i.e., 75 and 80 year expectations) (Covariate set 2).

There are a few important features of the subjective probabilities that
are worth discussing. First, as noted by @perozek2008 and others, about
2 percent of the respondents in the HRS provide invalid or impossible
responses to subjective mortality questions. For instance, some
respondents rate the probability that they will survive to 85 years old
greater than the probability that they will survive to 75. Researchers
have attributed this to a misunderstanding of the question and in
settings where life-tables are estimated, they cannot be used because
they are internally inconsistent. However, this is less of a concern in
a prediction setting where we are concerned about obtaining precise
$\hat Y$ values rather than correctly estimated $\beta$s or survivor
curves. In the cases where individual survival probabilities are
internally inconsistent, they will either contribute predictive power or
not to the resulting predictions.

\section{Data Cleaning and Processing}

Variable transformations and data processing are crucial to maximizing
predictive power. For instance, applying transformations to continuous
that bin values into new levels allows for greater non-parametric or
non-linear structure to be revealed, which in many cases improves
performance. To capture a richer structure from the available data, I
apply a series of feature engineering steps described below:

-   Create a new level indicator for factor variables, that have
    differing levels in the training and testing set.

-   Create a new level for factor variables called "unknown" to code
    them as missing (this will be used as a missing indicator later).

-   Impute numeric variables with the mean value of non-missing values
    and the mode for nominal variables.

-   Remove variables that have zero variance.

-   Convert all nominal variables to dummy variables.

After the pre-processing steps are applied to the data, the resulting
dataset comprises 17.2 thousand observations and 647 predictors. To
train the model, I randomly sample 80% of the data for the training
sample and 20% for the hold-out sample to evaluate predictions.

\section{Methods}

\subsection{Overview of Prediction Framework}

The goal of this paper is to predict longevity (death age in years)
among those who have died in the HRS sample with a secondary analysis
classifying individuals based on whether they lived to age 75. In this
section, I describe the preferred prediction model for the continuous
case. To predict longevity, I employ a variety of supervised machine
learning algorithms ranging from Ordinary Least Squares (OLS)
regression, to tree based methods (trees, forests, boosted trees) to
find the algorithm that best maps $death \ age$ as a function of
predictors $f(X)$.

For both longevity and survival to age 75, my preferred model is a
random forest. Tree-based methods generate predictions by partitioning
the predictor space (i.e., Xs) into non-overlapping groups and taking
the average value of death age in each group. The decision to split on a
variable is guided by maximizing a goodness of fit statistic (in this
case RMSE (described in \textbf{Asessing Goodness of Fit}) and is local
to each specific step in the decision tree. That is, the model builds
trees via the splitting objective to achieve the optimal improvement in
goodness of fit at each potential split. This is called recursive,
binary splitting; Rather than run a costly optimization routine
throughout an entire tree, the algorithm works in a stepwise fashion.
The random forest model builds on the simple regression tree in a few
important ways. First, random forests comprise multiple trees (hence the
forest) that are generated from bootstrap samples of the training data
that aide in reducing the variance in predictions from a single tree.
Second, at each split, random forests only consider a random subset of
predictors to split on, which helps de-correlate trees and avoid the
problem of the algorithm always selecting "highly predictive" features
to split on and thus undermining the advantage of averaging predictions
across multiple trees.

To generate predictions, I train a random forest model using 100 trees
and use five-fold cross-validation to tune the number of covariates
sampled for each split and the minimum number of nodes. For the
regression, I use RMSE as the loss function for death age prediction and
the log-loss for survival classification. The chosen values of the
tuning parameters for the main models predicting death age are reported
in Table 4 in the \textbf{Appendix}.

\subsection{Assessing Goodness of Fit}

To assess goodness of fit, I use a variety of measures for each outcome.
For death age, I use the RMSE and $R^2$. The two goodness of fit
measures are calculated as follows.First, RMSE is calculated as.

\begin{equation}
RMSE = \sqrt{\frac{1}{N} \sum_{i=1}^N (\hat Y_i - Y_i)^2}
\end{equation}

The RMSE measures the average of the deviations between predicted and
actual values. The RMSE can be naturally interpreted against the
standard deviation of the outcome variable (death age) as how well the
model does relative to simply using the average value to predict the
outcome. A lower RMSE value indicates better predictive performance.

Second, I use the $R^2$, which is calculated as follows.

\begin{equation}
 R^2 = 1 - \frac{\sum_{i=1}^{N} (Y_i - \hat{Y}_i)^2}{\sum_{i=1}^{N} (Y_i - \bar{Y})^2}
\end{equation}

The $R^2$ ranges from $- \infty$ to 1 and measures the share of the
variance explained by the model. A higher $R^2$ value indicates better
predictive performance.

For predicting survival past 75, I use the accuracy, sensitivity,
precision, and area under the curve (AUC). Each of these measures
targets a different aspect of model performance for binary outcomes. The
AUC is the focal and most holistic measure that captures the area under
the receiver-operator curve (ROC), which gives an overall assessment of
the model's ability to distinguish between classifications (i.e.,
survival to 75 versus dying before 75) rather than the performance of
classifying individuals into a certain category.

The accuracy of the model is the proportion of classifications, whether
positive or negative, that were correct. The precision is the proportion
of those that survived to 75 that were classified correctly out of true
and false positives. Finally, the sensitivity, or true positive rate,
refers to the proportion of the sample that survived past 75 classified
as such out of the total number of individuals who survived past 75.

\subsection{Interpreting Models}

A key limitation of machine learning methods is that they are often
"black boxes" where a researcher specifies a model and obtains
predictions, but which features contributed to those outcomes are either
not known or usually ignored. I address model interpretability in two
ways.

The first is built in to my prediction exercise by design. I contrast
two feature sets in all models. One contains 647 covariates spanning
health, education, wealth, income, social relationship quality, and
others captured in the HRS. The second contains all of the 647 variables
as well as a set of subjective health and mortality expectations
variables. Therefore, comparing the predictions from these two feature
sets both within and across models allows for an examination of variable
importance.

To further assess variable importance, I obtain highly important
features from random forest models and discuss which features contribute
strongly to predictions obtained across models.

\section{Results}

\subsection{Do Subjective Mortality Expectations Matter for Predicting Longevity?}

I first present analyses of the absolute and relative performance of
different machine learning methods for predicting longevity. Table 3
compares the performance of each method by covariate set, relative to a
baseline OLS regression of longevity on mortality expectations. Figure 1
displays results from Table 2 visually. First, a baseline regression of
longevity on mortality probabilities has an r-squared of $0.036$
indicating that the probabilities explain little of the variance in age
at death. The out of sample RMSE for this model is $9.116$, which
indicates that the model improves only slightly over using the mean
value of death age (RMSE = $9.283$).

In comparison to the baseline model, adding covariates to OLS improves
the RMSE by $36.992$ percent. Adding subjective probabilities in
addition to covariates improves the the RMSE by $37.065$, percent over
baseline. However, this relative improvement is small compared to the
covariate only model.

\begin{figure}[H]
  \centering 
  \includegraphics[width=\linewidth]{Images/long_improvement.jpeg}
  \caption{Absolute Improvement in Longevity Prediction}
  \end{figure}
\begin{figure}[H]
  \centering 
  \includegraphics[width=\linewidth]{Images/long_improvement_rel.jpeg}
  \caption{Relative Improvement in Longevity Prediction}
  \end{figure}

Predictive performance is similar with regularized models (LASSO, Ridge,
and Elastic Net) and single tree methods, both across covariate sets and
models. However, the largest increases in predictive performance are
obtained by using tree-based methods such as random forest and XGBoost.
In terms of predictive performance, the preferred model is a random
forest.

For the model with covariates only, the random forest improves
substantially on the baseline RMSE, by $47.962$ percent and achieves an
$R^2$ of $0.739$. Adding subjective probabilities provides essentially
no improvement in predictive power. However, in the XGBoost model,
despite performing worse overall than random forest, there are small
improvements.

Across quintiles of death age, the random forest model improves on
prediction for all quintiles. However, there are notable inequalities in
the magnitude of these improvements. The majority of the increase in
improvement is in the 3rd quintile, while the lowest improvement is
consistently in the 2nd quintile. Overall, the random forest model with
covariates and probabilities performs the best, but it is not
distinguishable from the covariate only random forest.

Next, I describe the secondary prediction exercise that seeks to predict
survival past 75 years of age using covariates and subjective
probabilities. Across models, subjective probabilities significantly
improve on the predictability of survival past 75 years old in all
models. Similarly, tree-based methods vastly outperform the traditional
regression and regularization methods for classifying individuals. The
preferred model for survival is an XGBoost model.

\begin{figure}[H]
  \centering 
  \includegraphics[width=\linewidth]{Images/survive_improvement.jpeg}
  \caption{Absolute Improvement in Survival Threshold Prediction}
  \end{figure}
\begin{figure}[H]
  \centering 
  \includegraphics[width=\linewidth]{Images/survive_improvement_rel.jpeg}
  \caption{Relative Improvement in Survival Threshold Prediction}
  \end{figure}

Using a logit model with subjective probabilities achieves a $0.768$
AUC, indicating that the model does better at classifying survival
relative to a random model that would have an AUC of $0.5$. Adding
covariates and covariates and probabilities improves the AUC by $8.606$
and $15.217$ percent, respectively. This suggests that even relative to
the naive, baseline model and covariates, the probabilities improve
prediction.

Similar to predicting death age, the regularized and single tree models
essentially perform the same as the LOGIT specification. The preferred
model, based on predictive performance, is the XGBoost model. This model
improves predictive performance by $21.359$ percent over baseline for
the covariate only model and $25.477$ percent for the model including
probabilities. In both absolute and relative terms, subjective
probabilities improve predictive power by about $4$ percentage points.
In contrast to predicting death age, subjective information appears to
matter for predicting survival.

\subsection{Variable Importance: Leading Alternative Predictors of Longevity}

In this section, I describe the top 10 most important features for
predicting longevity in the random forest model and survival to 75 in
the XGboost model.

Figure 5 displays the top 10 most important variables in the forest
according to the permutation importance metric. The permutation
performance metric is calculated by measuring how much the model's
performance drops in terms of goodness of fit with the exclusion of a
specific feature. Thus, highly important features that contribute the
most to goodness of fit receive higher rankings. Unsurprisingly for
predicting longevity, age deciles at different waves and the cumulative
number of waves that an individual is apart of over time contribute the
most to predicting longevity.

\begin{figure}[H]
  \centering 
  \includegraphics[width=\linewidth]{Images/vip_longevity.jpeg}
  \caption{Top 10 Relevant Predictors of Longevity in Preferred Random Forest Model}
  \end{figure}

Next Figure 6 displays the same feature importance metrics for the top
10 features in my XGBoost model predicting survival to age 75. Here the
most important features are quite different. Mixed family status (as
measured by whether the individual has biological and non-biological
children) appear quite important. Additionally, changes to self-rated
health, employment, and wealth are predictive as well. "p_live_80_LE",
the subjective probability that an individual survives past 80 years old
also predicts survival to 75 strongly. This is unsurprising since
individuals who have high subjective probabilities that they will
survive past 80, let alone 75 years, may also have important unobserved
information about wellbeing that is predictive of survival to 75.

\begin{figure}[H]
  \centering 
  \includegraphics[width=\linewidth]{Images/vip_survive.jpeg}
  \caption{Top 10 Relevant Predictors of Survival to Age 75 in Preferred XGBoost Model}
  \end{figure}

\section{Conclusion}

Do subjective mortality expectations matter for predicting longevity?
The answer is: it depends. Results from random forest models and several
other machine learning methods suggest that subjective probabilities
contribute little predictive information beyond what is obtained from
other covariates. However, for predicting survival past the age of 75,
the subjective mortality probabilities improve predictive power. These
results highlight the potential utility of incorporating subjective
measures into predictive models. Although subjective information
provides weakly predictive information about longevity, it can improve
the prediction of survival past thresholds.

\pagebreak
\section{Tables}
\begin{landscape}
\input{./Output/03_performance_long.tex}
\end{landscape}
\begin{landscape}
\input{./Output/04_performance_long.tex}
\end{landscape}

\section{Appendix}

````{=tex}
\begin{landscape}

```{r Tuning Parameter Table}

# make tuning parameter table 
data.frame(
  # Model tuning Parameters
  Algorithm = c("OLS",
                "Lasso",
                "Lasso",
                "Ridge",
                "Ridge",
                "Elastic Net",
                "Elastic Net",
                "Tree",
                "Tree",
                "Random Forest",
                "Random Forest",
                "XGBoost",
                "XGBoost"),
 `Feature Set` = c("-", 
                   rep(c("Controls",
                         "Controls + Probs"),6)),
 `N Predictors` = c(647,460,483,
                    rep(547,2),
                    473,487,
                    rep(c(647),6)),
  Mixture =     c("-",
                  "1",
                  "1",
                  "0",
                  "0",
                  "0.5",
                  "0.75",
                  rep("-",6)),
  Lambda =      c("-",
                  "0",
                  "0",
                  "0",
                  "0",
                  "0",
                  "0",
                  rep("-",6)),
  `M Try` =      c(rep("-",7),
                  435,435,
                  145,151,
                  "-","-"
                  
                   ),
  `Min Nodes` =  c(rep("-",7),
  40,40,
  30,30,
  40,40),
  `N Trees` =    c(rep("-",7),
                   1,1,rep(100,4)),
  `Tree depth` = c(rep("-",11),8,8),
  `Learn Rate` = c(rep("-",11),rep(.3,2)),
  `Sample size` = c(rep("-",11),rep(1,2)),
  # Cross-validation 
  `CV Folds` = c("-",rep(5,12)),
 check.names = F
) %>% 
  datasummary_df(
    output = "kableExtra",
    title = "Tuning Paremeters by Model (Longevity)",
    notes = "All parameters are tuned using cross-validation except learn rate, number of trees, and number of folds."
  ) %>% 
  column_spec(1:2,width = "1in") %>%
  column_spec(3:11,width = ".5in")

```

\end{landscape}
````

\section{References}
