---
title: | 
  | Do Subjective Mortality Expectations Matter for Predicting Longevity?
author: |
    | Thelonious Goerz 
    | Department of Sociology 
    | PUBPOL 5390
date: "December 2, 2025"
output: 
  beamer_presentation:
    slide_level: 2
    theme: "default"
    colortheme: "dove"
header-includes: 
  - \usepackage{framed} 
  - \usepackage{amsmath}
  - \usepackage{tcolorbox} 
  - \usepackage{pdflscape}
  - \newcommand{\blandscape}{\begin{landscape}}
  - \newcommand{\elandscape}{\end{landscape}}
  - \usepackage{multirow} 
  - \usepackage{xcolor} 
  - \usepackage{tabularray}
  - \def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}} 
  - \usepackage{setspace} 
  - \usepackage{threeparttable}
  - \usepackage{booktabs}
  - \usepackage{siunitx}
  - \usepackage{tikz}
  - \usepackage{tabularray}
  - \usepackage{float}
  - \usepackage{graphicx}
  - \usepackage{rotating}
  - \usepackage[normalem]{ulem}
  - \UseTblrLibrary{booktabs}
  - \UseTblrLibrary{siunitx}
  - \newcommand{\tinytableTabularrayUnderline}[1]{\underline{#1}}
  - \newcommand{\tinytableTabularrayStrikeout}[1]{\sout{#1}}
  - \NewTableCommand{\tinytableDefineColor}[3]{\definecolor{#1}{#2}{#3}}
classoption: "aspectratio=169"
link-citations: yes
linkcolor: blue
font_size: 11pt
bibliography: references.bib
---

```{r setup, include=TRUE, results="hide", echo = FALSE, message = F}

 rm(list = objects())
 # THIS IS THE WIDTH OF THE CODE                                                ###
 knitr::opts_chunk$set(echo = F,
                       resuls = T, message = F, warning = F)
 
 # knitr::opts_chunk$set(fig.width = unit(16,"cm"), fig.height = unit(9,"cm"))
 library(tidyverse)
 library(modelsummary)
 library(kableExtra)
 
```

## Twin Motivations: (1) Policy

\begin{columns}  % T aligns the tops
    % Left side (text)
    \column{0.6\textwidth}
    
-   The population is aging, with some of the highest growth occurring for those over 65+

-   Predicting longevity is important for population health research because it provides crucial input information to guide current and future decisions about how to care and support an aging population

  \column{0.4\textwidth}
  \begin{flushright}
  \begin{figure}[H]
  \centering 
  \includegraphics[width=\linewidth]{Images/mpidr_paper.jpeg}
  \caption{Badolato et al. 2023}
\end{figure}
  \end{flushright}
\end{columns}

## Twin Motivations: (2) Substantive

-   Sociology [@merton1948; @vaisey2009], and economics [@velásquez-giraldo; @bénabou2016] recognize the theoretically predictive power of individuals' beliefs (e.g., financial decisions, health, etc.)

-   Subjective beliefs about survival may matter for predicting longevity in prediction models

## Background

-   Past work shows that predicting longevity is challenging and highly uncertain [@breen_demographic_2023; @badolato2023]

-   Moreover, correlations between individual mortality expectations and death age are weak, but their predictive power remains unclear

-   There are also stark inequalities in predictability across groups (e.g., race, education, gender)

    -   Subjective information may help

## Present Study

-   Idea: Lets bring these ideas to bear with a prediction exercise 

-   Goal: Predict longevity (death age) and survival past 75 years of age

-   Evaluate the predictability using (1) Covariate feature set, (2) Covariates + Subjective Probabiltiies

-   Conceptually, we think of the subjective probabilities as "utility" features -- they enrich predictions made with other, more powerful, covariates

## Data

-   Health and Retirement Study

-   Panel dataset of 32k individuals representative of US population 50+, (170,000 person-observations)

-   Rich feature set including health, wealth, education, work, stocks, social network information, caregiving

## Machine Learning Framework

```{r}
data.frame(
  Outcome = c("Longevity","Survival to 75"),
  Models = c(
    "OLS, LASSO, Ridge, Elastic Net, Tree, Random Forest, XGBoost",
     "Logit, LASSO, Ridge, Elastic Net, Classification Tree, Random Forest, XGBoost"),
  Evaluation = c("RMSE, $R^2$",
                 "Accuracy, Precision, Sensitivity, AUC")
) %>% 
  datasummary_df(
    title = "Summary of Machine Learning Models by Outcome",
    output = "kableExtra"
  ) %>% 
    column_spec(1:3, width = "1.5in")

```

## Data Processing and Tuning Model Parameters

-   For each model:

    -   Reshape data to wide format
    -   Filter to sample of those who have died (17.2k)
    -   Impute missing predictors (mean,mode) and create missing indicators [@mullainathan2017]
    -   Recode all factor and character variables to one-hot indicators

-   Use 5-fold cross-validation to tune parameters

    -   All parameters are tuned except for:

        -   Learn rate (XGboost)

        -   N Trees (Tree, Random Forest, XGBoost)

## Main Results: Absolute Improvement in Longevity

  \begin{figure}[H]
  \centering 
  \includegraphics[width=\linewidth,height=.7\textheight]{Images/long_improvement.jpeg}
  \caption{Absolute Improvement in Longevity Prediction}
  \end{figure}
  
## Main Results: Relative Improvement in Longevity

  \begin{figure}[H]
  \centering 
  \includegraphics[width=\linewidth,height=.7\textheight]{Images/long_improvement_rel.jpeg}
  \caption{Relative Improvement in Longevity Prediction}
  \end{figure}
  
## Main Results: Relative Improvement in Longevity (Quintile)

  \begin{figure}[H]
  \centering 
  \includegraphics[width=\linewidth,height=.7\textheight]{Images/long_improvement_q.jpeg}
  \caption{Relative Improvement in Longevity Prediction by Quintile of Death Age (Best Fitting Models)}
  \end{figure}


## Main Results: Absolute Improvement in Survival to 75

  \begin{figure}[H]
  \centering 
  \includegraphics[width=\linewidth,height=.7\textheight]{Images/survive_improvement.jpeg}
  \caption{Absolute Improvement in Survival Threshold Prediction}
  \end{figure}
  
## Main Results: Relative Improvement in Survival to 75

  \begin{figure}[H]
  \centering 
  \includegraphics[width=\linewidth,height=.7\textheight]{Images/survive_improvement_rel.jpeg}
  \caption{Relative Improvement in Survival Threshold Prediction}
  \end{figure}

## Main Results: Survival to 75 (Calibration Plot)

  \begin{figure}[H]
  \centering 
  \includegraphics[width=\linewidth,height=.7\textheight]{Images/survive_calib.jpeg}
  \caption{Calibration Plot of Full Feature Set XGBoost Algorithm Comparing Predicted Survival Probabilities and Actual Shares}
  \end{figure}
  
## Conclusion

- Tree-based methods with rich covariates improve on the baseline predictability of longevity and survival substantially (50$\%$ and 300$\%$ respectively)

- Across models, subjective mortality expectations do seem to aid in prediction, but only in small magnitudes and among certain groups

- Future efforts to collect data on subjective mortality could explore new ways of asking the questions (probabilites are hard for people to think about)







## References
\small
