---
title: "Lecture 3"
author: "Thelonious Goerz"
date: "2025-09-02"
output: pdf_document
---

**Goodness of Fit Statistics**

-   MSE and RMSE

    -   Population expectation

    -   within-sample calculation

    -   out of sample calculation

Sum of squared errors - what we are estimating

$$
OOS \ MSE = \frac 1 n \sum (y-\hat y)^2 
$$

The goodness of fit statistics are trained on the sample to fit that sample, so it is "overfit" in a sense because it it is trying to predict the data it has. What we additionally want to know is how this performs out of sample.

We want to learn about some population expectation (what theoretically we are trying to target)

$$
Pop = E \big [\sum (y-\hat y)^2 \big]
$$

This is the in sample sum of squares with a degrees of freedom correction that we have for the sample.

$$
In \ Sample = \frac 1 {n-k} \sum (y-\hat y)^2
$$

Measured residuals will be systematically too small compared to the sum of true model errors, it will be downward biased due to variation. The n-k correction helps account for this the ratio. Within-sample GOF is biased, so best to get out of sample which will beadjusted in a sense.

-   RMSE is more interpretable, and has an analogue to size of errors

## R-squared 

$$
SSTotal = \sum (y-\hat y)^2 \\
SSExp = \sum(\hat y - \bar y)^2 \\
R^2 = \frac {SSExp} {SSTotal} \\
= 1- \frac {SSExp} {SSTotal} \\
SSErr = \sum (y-\hat y)^2 
$$

## RMSE 

-   What is a good RMSE?

    -   Smaller is better

    -   Interpret relative to a model with just the intercept

# In Class Application 

```{r}
rm(list = ls())
library(readr)
library(tidyverse)
library(fixest)
library(Metrics)
library(magrittr)
data_a = read_csv("../Data/births_for_class_A.csv")
data_b = read_csv("../Data/births_for_class_B.csv")
data_c = read_csv("../Data/births_for_class_C.csv")
data_d = read_csv("../Data/births_for_class_D.csv")
```
```{r}
data_a %<>% mutate(gestation_weeks = factor(gestation_weeks)) %>% filter(!is.na(gestation_weeks)) %>% droplevels() %>% na.omit()
data_b %<>% mutate(gestation_weeks = factor(gestation_weeks)) %>% filter(!is.na(gestation_weeks)) %>% droplevels() %>% na.omit()
data_c %<>% mutate(gestation_weeks = factor(gestation_weeks)) %>% filter(!is.na(gestation_weeks)) %>% droplevels() %>% na.omit()
data_d %<>% mutate(gestation_weeks = factor(gestation_weeks)) %>% filter(!is.na(gestation_weeks)) %>% droplevels() %>% na.omit()
```

```{r Model Train}

model <- lm(birth_weight ~ as.factor(month) + as.factor(dayofweek) + time + in_hospital + poly(mom_age, 2) + as.factor(mom_race_hispanic_status)*as.factor(dad_race_hispanic_status) + mom_bmi + as.factor(residence_status) + poly(dad_age, 2) + prior_other_terminations + prior_births_dead +  number_prenatal_care_visits + num_cigarattes_before_pregnancy + pregnancy_weight_gain + gestation_weeks+  as.factor(mom_educ) * as.factor(dad_educ),
            data = data_a)


summary(model)


```

```{r Test}
predictions = predict(model, newdata = data_b)

levels(data_a$gestation_weeks)

rmse_test = rmse(data_b$birth_weight,predictions)

## R-squared 
obs <- data_b$birth_weight

# Compute R-squared manually
rss <- sum((obs - predictions)^2)       
tss <- sum((obs - mean(obs))^2) 
rsq <- 1 - rss/tss

data.frame(R2 =rsq,
           RMSE = rmse_test) %>% 
  modelsummary::datasummary_df()

```

What to do about the new gestation level 