---
title: "Lecture 11"
author: "Thelonious Goerz"
date: "2025-10-09"
output: html_document
---

# Notes

Binary dependent variables - classification

-   Leads to two different conceptual answers

    -   classify which category something is in

    -   for each category, what is the probability that the person that is in each category

-   For Binary DV

    -   different ways to interpret the models

    -   different ways to assess goodness of fit

-   LPM

    -   Regression is an approximation of the conditional expectation function

    -   In probability theory, when Y is discrete (0,1) the conditional expectation function is the prob that Y=1.

    -   In regression, change in x gives change in probability of y.

    -   Can lead to unrealistic predictions, but error distribution is strange. Error can only take 0,1. Variance of those errors depends on x values. Creates heteroskedacticity issues.

-   Logit and probit

    -   We want a model that is constrained to 0,1 and we want changes to reflect nonlinearities (decreasing marginal returns)

    -   Take Xs as index and turn them into Pr(Y=1) (ODDS)

    -   Choose a distribution function

        -   logistic or normal CDF (logit, probit)

# In class exercise

```{r}
rm(list = ls())
library(tidymodels)
library(readr)
library(tidyverse)
library(magrittr)
```

```{r Read Data}
g = read_csv("../Data/births_for_class_Gsmall.csv")
h = read_csv("../Data/births_for_class_H.csv")
```

```{r Store Results}
set.seed(999)
performance_df = 
  data.frame(
    Model = c("OLS","LASSO","RIDGE","ELASTIC NET"),
    `SD (Birthweight) In-sample` = c(rep(sd(g$birth_weight),4)),
    `SD (Birthweight) Out of Sample` = c(rep(sd(h$birth_weight),4)),
    `RMSE IS` = c(rep(NA_real_,4)),
    `RMSE CV` = c(rep(NA_real_,4)),
    `RMSE OOS` = c(rep(NA_real_,4)),
    lambda = c(rep(NA_real_,4)),
    alpha = c(rep(NA_real_,4))
  )

performance_df

```

```{r Overfit OLS}

g_base = g
h_base = h

ols = linear_reg(
  mode = "regression",
  engine = "lm"
) 

# Create Feature Set 
g %<>% 
  mutate(across(c(gestation_weeks,mom_age,in_hospital,mom_educ,residence_status,mom_height_inches), as.factor)) 
h %<>% 
  mutate(across(c(gestation_weeks,mom_age,in_hospital,mom_educ,residence_status,mom_height_inches), as.factor)) 

bw_recipe = 
  recipe(birth_weight ~ ., data = g) %>%
  step_dummy(all_nominal_predictors(),one_hot = T)

bw_recipe_oos = 
  recipe(birth_weight ~ ., data = h) %>% 
  step_dummy(all_nominal_predictors())


# workflow 
wflow = 
 workflow() %>% 
  add_model(ols) %>% 
  add_recipe(bw_recipe)

# Fit Model 
fit_ols =
  wflow %>% 
  fit(data = g)

# fit on OOS # 
wflow_oos = 
 workflow() %>% 
  add_model(ols) %>% 
  add_recipe(bw_recipe_oos)

fit_oos = 
  wflow_oos %>% 
  fit(data = h)

g_base$y_hat_ols_is = predict(fit_ols,new_data = g)$.pred
h_base$y_hat_ols_os = predict(fit_ols,new_data = h)$.pred

# Cross-validate RMSE 
folds = vfold_cv(g,v = 10)
cv_rmse_ols= tune_grid(ols, bw_recipe, resamples = folds)

# Calculate RMSE # 
ols_is_rmse = rmse(data = g_base,
     truth = birth_weight,
     estimate = y_hat_ols_is)

ols_oos_rmse = rmse(data = h_base,
     truth = birth_weight,
     estimate = y_hat_ols_os)

# Save Metrics 
performance_df[performance_df$Model == "OLS","RMSE.IS"] = ols_is_rmse$.estimate
performance_df[performance_df$Model == "OLS","RMSE.OOS"] = ols_oos_rmse$.estimate
performance_df[performance_df$Model == "OLS","RMSE.CV"] = show_best(cv_rmse_ols,metric = "rmse")$mean

performance_df

```

```{r LASSO}
set.seed(999)
# #####
# Tune Lasso parameters 

# ####

lasso_tune = linear_reg(
  mode = "regression",
  engine = "glmnet",
  mixture = 1, 
  penalty = tune()
) 

# Cross-validate lambda # 
cv_lambda = tune_grid(lasso_tune, 
                      bw_recipe, 
                      resamples = folds, 
                      tune = penalty,
                      grid = expand.grid(penalty = 0:100))

# Pick lambda # 

best_lambda_lasso = show_best(cv_lambda)[1,"penalty"]

# Declare Model # 
lasso = linear_reg(
  mode = "regression",
  engine = "glmnet",
  mixture = 1, 
  penalty = as.numeric(best_lambda_lasso)
) 

# Estimate Model in-sample # 

fit_lasso = workflow() %>% 
  add_model(lasso) %>% 
  add_recipe(bw_recipe) %>%
# Fit Model 
  fit(data = g)

# fit on OOS # 

fit_oos =  workflow() %>% 
  add_model(lasso) %>% 
  add_recipe(bw_recipe_oos) %>%
  fit(data = h)


g_base$y_hat_lasso_is = predict(fit_lasso,new_data = g)$.pred
h_base$y_hat_lasso_os = predict(fit_lasso,new_data = h)$.pred

# Cross-validate RMSE 
# folds = vfold_cv(g,v = 10) Folds defined earlier
cv_rmse_lasso= tune_grid(lasso, bw_recipe, resamples = folds)

# Calculate RMSE # 
lasso_is_rmse = rmse(data = g_base,
     truth = birth_weight,
     estimate = y_hat_lasso_is)

lasso_oos_rmse = rmse(data = h_base,
     truth = birth_weight,
     estimate = y_hat_lasso_os)

# Save Metrics 
performance_df[performance_df$Model == "LASSO","RMSE.IS"] =  lasso_is_rmse$.estimate
performance_df[performance_df$Model == "LASSO","RMSE.OOS"] = lasso_oos_rmse$.estimate
performance_df[performance_df$Model == "LASSO","RMSE.CV"] = show_best(cv_rmse_lasso,metric = "rmse")$mean
performance_df[performance_df$Model == "LASSO","lambda"] = as.numeric(best_lambda_lasso)
performance_df[performance_df$Model == "LASSO","alpha"] = 1

# Create cross-validation plot of lambda #

cv_plot = collect_metrics(cv_lambda) %>% 
  filter(.metric == "rmse") %>% 
  ggplot(aes(penalty,mean)) + 
  geom_line(lwd =2, color = "blue") + 
  geom_vline(xintercept = as.numeric(best_lambda_lasso),color = "red",lwd = 1.5) + 
  geom_hline(yintercept = performance_df$SD..Birthweight..Out.of.Sample[1],color = "grey50",lwd = 1.5,linetype = "dashed") + 
  theme_minimal() + 
  labs(x = "Lambda",
       y = "RMSE",
       caption = "Red Line is CV best lambda. Grey line is OOS SD(Birthweight).",
       title = "Cross-Validataion lambda Plot (LASSO)")

ggsave(cv_plot,
       file = "../HW/In_class_Exercise/cv_plot_d14.jpeg")

```

```{r Ridge}
# #####
# Tune Ridge Parameters
# ####
ridge_tune = linear_reg(
  mode = "regression",
  engine = "glmnet",
  mixture = 0, 
  penalty = tune()
) 

# Cross-validate lambda # 
cv_lambda2 = tune_grid(ridge_tune, 
                      bw_recipe, 
                      resamples = folds, 
                      tune = penalty,
                      grid = expand.grid(penalty = 0:100))

# Pick lambda # 

best_lambda_ridge = show_best(cv_lambda2)[1,"penalty"]

# Declare Model # 
ridge = linear_reg(
  mode = "regression",
  engine = "glmnet",
  mixture = 0, 
  penalty = as.numeric(best_lambda_ridge)
) 

# Estimate Model in-sample # 

fit_ridge = workflow() %>% 
  add_model(ridge) %>% 
  add_recipe(bw_recipe) %>%
# Fit Model 
  fit(data = g)

g_base$y_hat_ridge_is = predict(fit_ridge,new_data = g)$.pred
h_base$y_hat_ridge_os = predict(fit_ridge,new_data = h)$.pred

# Cross-validate RMSE 
# folds = vfold_cv(g,v = 10) Folds defined earlier
cv_rmse_ridge= tune_grid(ridge, bw_recipe, resamples = folds)

# Calculate RMSE # 
ridge_is_rmse = rmse(data = g_base,
     truth = birth_weight,
     estimate = y_hat_ridge_is)

ridge_oos_rmse = rmse(data = h_base,
     truth = birth_weight,
     estimate = y_hat_ridge_os)

# Save Metrics 
performance_df[performance_df$Model == "RIDGE","RMSE.IS"] =  ridge_is_rmse$.estimate
performance_df[performance_df$Model == "RIDGE","RMSE.OOS"] = ridge_oos_rmse$.estimate
performance_df[performance_df$Model == "RIDGE","RMSE.CV"] = show_best(cv_rmse_ridge,metric = "rmse")$mean
performance_df[performance_df$Model == "RIDGE","lambda"] = as.numeric(best_lambda_ridge)
performance_df[performance_df$Model == "RIDGE","alpha"] = 1

```

```{r Elastic Net}
# #####
# Tune Parameters
# ####
elasticnet_tune = linear_reg(
  mode = "regression",
  engine = "glmnet",
  mixture = tune(), 
  penalty = tune()
) 

# Cross-validate lambda # 
cv_en = tune_grid(elasticnet_tune, 
                      bw_recipe, 
                      resamples = folds, 
                      tune = c(penalty,mixture),
                      grid = expand.grid(penalty = 0:100,
                                         mixture = c(.25,.5,.75)))

# Pick lambda # 

best_penalty_en = show_best(cv_en)[1,"penalty"]
best_mixture_en = show_best(cv_en)[1,"mixture"]

# Declare Model # 
en = linear_reg(
  mode = "regression",
  engine = "glmnet",
  mixture = .5,
  penalty = as.numeric(best_penalty_en)
) 

# Estimate Model in-sample # 

fit_en = workflow() %>% 
  add_model(en) %>% 
  add_recipe(bw_recipe) %>%
# Fit Model 
  fit(data = g)

g_base$y_hat_en_is = predict(fit_en,new_data = g)$.pred
h_base$y_hat_en_os = predict(fit_en,new_data = h)$.pred

# Cross-validate RMSE 
# folds = vfold_cv(g,v = 10) Folds defined earlier
cv_rmse_en= tune_grid(en, bw_recipe, resamples = folds)

# Calculate RMSE # 
en_is_rmse = rmse(data = g_base,
     truth = birth_weight,
     estimate = y_hat_en_is)

en_oos_rmse = rmse(data = h_base,
     truth = birth_weight,
     estimate = y_hat_en_os)

# Save Metrics 
performance_df[performance_df$Model == "ELASTIC NET","RMSE.IS"] =  en_is_rmse$.estimate
performance_df[performance_df$Model == "ELASTIC NET","RMSE.OOS"] = en_oos_rmse$.estimate
performance_df[performance_df$Model == "ELASTIC NET","RMSE.CV"] = show_best(cv_rmse_en,metric = "rmse")$mean
performance_df[performance_df$Model == "ELASTIC NET","lambda"] = as.numeric(best_penalty_en)
performance_df[performance_df$Model == "ELASTIC NET","alpha"] = as.numeric(best_mixture_en)

```

```{r}
performance_df

bw_recipe$
```

