---
title: "Lecture 16"
author: "Thelonious Goerz"
date: "2025-10-30"
output: html_document
---

# Ensembles of trees

-   How many trees

    -   more is better - averaging reduces noise

    -   diminishing returns

    -   only cost is commutation time

    -   fold wisdom - 50-500

-   Ensemble solves the noise

# Bagging

Regular BS - oos

implication is that about 1/3 observations are not used

if sample without replacement, selecting a fraction \<1 of observations

The observations not used are like a mini testing sample

so the gof on those observations gives a "free" CV estimate

-   Out of Bag goodness of fit estimate (averaging across trees)

    -   Can be thought of an out of sample Goodness of fit

-   Random forests

    -   baggins is not good enough

    -   the trees that we get from bagging may be too correlated

        -   if there are very extra predictive Xs (e.g., Gestation) then the dominant feature will be used in the trees too much

    -   Solution: de-correlate the trees

    -   only consider a subset of features at each node of each tree,

    -   A random fraction of features can be used at each node (e.g., within a tree at each node you look at only a random subset)

-   What fraction?

    -   folk wisdom (textbook) answer

    -   regression trees = k/3

    -   classification trees $\sqrt K$

    -   this can think of this as a tuning parameter

-   **Athey:** trees are self-regularizing

    -   built into the procedure – not a post-estimation procedure like in other methods

-   variable importance measures

    -   what variables matter? in a single tree we can look at what its splitting on

        -   but in a forest, it is very black box/

        -   but there are stats for this - look at each tree and look at each node (count of times that variables were chosen)

        -   We can also look at all splits and get a measure of how much that feature improves MSE - average across all

-   Classification trees

    -   predictions for binary outcomes

    -   $\hat p_{mk}$ predictions for outcome K, leaf M

-   GOF = 3 choices built on pred prob and majority vote

    -   accuracy

    -   gini

    -   cross entropy

    -   node impurity - goal is a low value for each outcome

```{r}
library(tidymodels)
library(here)
library(readr)

d = read_csv(here("Data","births_for_class_G.csv"))
```

```{r}
t = initial_split(d, prop = 3/4)

train = training(t)
test = testing(t)


tree_rec <- 
  recipe(birth_weight ~ ., data = train) %>%
  step_dummy(all_nominal(), -all_outcomes()) 

tree_prep <- prep(tree_rec)
juiced <- juice(tree_prep)

```

```{r}
tune_spec <- rand_forest(
  mtry = tune(),
  trees = 1000,
  min_n = tune()
) %>%
  set_mode("regression") %>%
  set_engine("ranger")

tune_wf <- workflow() %>%
  add_recipe(tree_rec) %>%
  add_model(tune_spec)

```

```{r}
set.seed(234)
trees_folds <- vfold_cv(train)

doParallel::registerDoParallel()

set.seed(345)
tune_res <- tune_grid(
  tune_wf,
  resamples = trees_folds,
  grid = 20
)

tune_res

```
