---
title: " 5390 Lecture 2"
author: "Thelonious Goerz"
date: "2025-08-28"
output: html_document
---

**Goals of data analysis**

-   Why predict $\hat{y}$ ?

    -   predict to get potential players or make decisions

    -   policy question

    -   forecasting

-   Policy Prediction Problems

-   Prediction problems are embedded within causal questions (e.g. selection on onbservables)

-   Variation in treatment effect using machine learning tools

**OLS**

-   Xs are features

-   Y = label

-   Betas are parameters

-   GOF

    -   r-squared

    -   RMSE

    -   SER

    -   MAPE

-   Standard deviation of the Y is the RMSE for a model with no predictors

-   

# Prediction Exercise

```{r}
library(readr)
library(tidyverse)
library(fixest)
data = read_csv("../Data/births_for_class.csv")


```

```{r}
# Descriptive Context 
data %>% 
  summarise(N = n(), 
            Mean = mean(birth_weight),
            min = min(birth_weight),
            max = max(birth_weight))

data %>% 
  group_by(gestation_weeks) %>%
  summarise(N = n(), 
            Mean = mean(birth_weight),
            min = min(birth_weight),
            max = max(birth_weight))
```

```{r}
# model 

m = lm(birth_weight~gestation_weeks*mom_race_hispanic_status + mom_educ + mom_age + in_hospital, data = data)

summary(m)

data$in_hospital


```
