---
title: "Lecture 14"
author: "Thelonious Goerz"
date: "2025-10-23"
output: html_document
---

# Tree-Based Methods

-   Can add regularization parameters to tree-based methods

-   Classification tree or regression tree

-   tree building process

    -   splitting rule

    -   goal - partition the set of Xs into a collection of rectangular regions

        -   each observation goes into one region

        -   predicted value is the same for all observations in the same region

        -   continuous: ybar

        -   binary: probability

-   idea - partitioning the parameter space into groups

-   Building a regression tree

    -   Start with a single tree within training data

    -   calculate y_hat = y_bar for data

    -   compute MSE for single pooled note: Var(Y)

    -   start by splitting on a variable X_j (years of experience)

        -   Consider a cutoff value C_i (cutoff is 5.5)

    -   Get candidate splits - calculate Y_hat = ybar within each split

    -   Get new MSE

# In class exercise

```{r}
rm(list = ls())
library(tidymodels)
library(readr)
library(tidyverse)
library(magrittr)
```

```{r Read Data}
g = read_csv("../Data/births_for_class_Gsmall.csv")
h = read_csv("../Data/births_for_class_H.csv")
```

```{r DATA PREP}
g %<>% mutate(label = "G")
h %<>% mutate(label = "H")
c = rbind(g,h)

g$birth_weight
## Create LHS ## 
c %<>% 
  mutate(LBW = factor(ifelse(birth_weight < 2500,"low","normal"))) %>% 
  mutate(across(c(gestation_weeks,in_hospital,mom_educ,residence_status), as.factor)) 

G = c[c$label == "G",] %>% select(-label)
H = c[c$label == "H",] %>% select(-label) %>% filter(gestation_weeks %in% G$gestation_weeks)


```

```{r Estimate Probit}

binary = logistic_reg() %>%
  set_engine("glm", family = binomial(link = "probit")) 

bw_recipe = 
  recipe(LBW ~ ., data = G)

bw_recipe_oos = 
  recipe(LBW ~ ., data = H) 


# workflow 
wflow = 
 workflow() %>% 
  add_model(binary) %>% 
  add_recipe(bw_recipe)

# Fit Model 
fit_probit =
  wflow %>% 
  fit(data = G)

# fit on OOS # 
wflow_oos = 
 workflow() %>% 
  add_model(binary) %>% 
  add_recipe(bw_recipe_oos)

fit_oos = 
  wflow_oos %>% 
  fit(data = H)

G$y_hat_probit_is = predict(fit_probit,new_data = G,type = "prob")$.pred_low
H$y_hat_probit_os = predict(fit_oos,new_data = H,type = "prob")$.pred_low

```

```{r Confusion Matrix Train}

augment(fit_probit, new_data = G) |>
  conf_mat(truth = LBW, estimate = .pred_class)

```

```{r Confusion Test}

augment(fit_probit, new_data = H) |>
  conf_mat(truth = LBW, estimate = .pred_class)


# accuracy
(631+ 10365) / nrow(H)
# tpr
631 / (631+143)
# FPR
326 /(326 + 10365)

```

```{r}
fit
augment(fit_probit, new_data = G) %>%
roc_curve(truth = LBW,
          .pred_low) %>%
  autoplot()


augment(fit_probit, new_data = H) %>%
roc_auc(truth = LBW,
          .pred_low) %>% summary()
```


```{r Calibration Plot}
library(KernSmooth)
# Get quantiles
q = quantile(H$y_hat_probit_os,probs = c(.05,.95))

H %<>% mutate(low_bw = ifelse(LBW=="low",1,0))
  
calibration = locpoly(x =H$y_hat_probit_os,y = H$low_bw , bandwidth = .5,
               degree = 3,
               kernel = "Epanechnikov",
               range.x = c(q[1],q[2]),gridsize = 1000)

calibration %<>% data.frame() 


ggplot() + 
  geom_line( size = 2, color = "blue",aes(calibration$x,calibration$y)) + 
  theme_bw() 

```









