---
title: "Lecture 17"
author: "Thelonious Goerz"
date: "2025-11-04"
output: html_document
---

# Project Proposal

-   Slide deck and write up are due on the 17th

-   Sufficiently address each point

------------------------------------------------------------------------

# Boosting

-   Build a Tree

-   Get Residuals

-   Build a second tree and predict the residuals

-   new prediction: sum of the trees (tree 1 + tree 2)

-   Repeat

-   Each tree tries to fix the mistakes of the last version of the model

-   A series of trees 1-B

-   For each step: $\hat b^b (x)$ function to predict residuals from tree b as a function of X

-   often each tree is simple (stump = 1 split)

-   Prediction at stage B = $\hat f(x) = \sum_b^B \hat f^b (X)$

-   Boosting shrinkage

    -   idea - don't update the model fully, just take a step of size $\lambda$ in the direction of the new tree.

    -   Lambda is a shrinkage parameter but not the same as LASSO In this case we are shrinking the updates to the predictions.

    -   Similar steps =\> maybe better than many-tree performance

    -   Lambda tune $\hat f(x) = \sum_b^B \times \lambda \hat f^b (X)$

-   

```{r}

library(tidymodels)
library(here)
library(readr)
library(vip)
library(tictoc)
library(ranger)
library(solitude)
G = read_csv(here("Data","births_for_class_G.csv"))
H= read_csv(here("Data","births_for_class_H.csv"))
```

```{r}
d = rbind(G,H)
```

```{r}
set.seed(109900)

d = d %>% select(-gestation_weeks)
t = initial_split(d, prop = 3/4)

train = training(t)
test = testing(t)

tree_rec <- 
  recipe(birth_weight ~., data = train) 

tree_prep <- prep(tree_rec)
juiced <- juice(tree_prep)

```

```{r Tuning}

tune_spec <- rand_forest(
  mtry = tune(),
  trees = 100,
  min_n = 3
) %>%
  set_mode("regression") %>%
  set_engine("ranger",
             importance = "impurity"
             )

tune_wf <- workflow() %>%
  add_recipe(tree_rec) %>%
  add_model(tune_spec)

```

```{r}
tic()
set.seed(234)
trees_folds <- vfold_cv(train,v = 5)

param_grid <- grid_regular(mtry(range = c(0, 10)), levels = 5)

doParallel::registerDoParallel()

tune_res <- tune_grid(
  tune_wf,
  resamples = trees_folds,
  grid = param_grid
)
toc()


collect_metrics(tune_res)
```

```{r}
best_complexity <- select_best(tune_res, metric = "rmse")

rand_forest_final <- finalize_workflow(tune_wf, best_complexity)

rf_fit <- fit(rand_forest_final, data = train)

train_gof = data.frame(
truth = as.numeric(rf_fit$pre$mold$outcomes$birth_weight),
     estimate = as.numeric(predict(rf_fit,train)$.pred))
rmse(data = train_gof,
     truth = truth, 
     estimate = estimate)

```

```{r}
sqrt(267062.1 )

pred = predict(rf_fit, new_data = test)

test$pred = pred$.pred

rmse(data = test,
     truth = birth_weight,
     estimate = pred) 

sd(train$birth_weight)

# variable importance
rf_fit %>%
extract_fit_parsnip() %>%
      vip(num_features = 10) 

 ranger_obj <- rf_fit %>%
      extract_fit_engine()
 
ranger_obj <- rf_fit %>% extract_fit_engine()

terminalNodesDepth(ranger_obj) %>%
  summarise(max(depth),
            mean(depth))

```

How to get tree depth trained? How to set tree depth

```{r}
# OLS 

ols_model = 
  linear_reg(mode = "regression",
             engine =  "lm")

ols_wf <- workflow() %>%
  add_recipe(tree_rec) %>%
  add_model(ols_model)

ols_fit = fit(ols_wf,data = train)

augment(ols_fit,new_data = test) %>% 
  rmse(truth = birth_weight, 
       estimate = .pred)
```
